---
title: "Untitled"
output: html_document
date: "2026-03-16"
---

# Load package and data  

```{r}
library(tidyverse)
library(sbm)
library(phyloseq)
library(viridis)
library(alluvial)
```


```{r}
load(file = "outputs/heavy_computation_save/subsample_list_bact.Rdata")
load(file = "outputs/heavy_computation_save/subsample_sbm_bact.Rdata")
load(file = "outputs/clean_data_to_analise/ps_bacteria.Rdata")
```

# Arrange some of the main data

```{r}
ps_bacteria_grid <- merge_samples(ps_bacteria, "grid")
ps_bacteria_grid %>%
  otu_table() %>% 
  t() %>%
  as.data.frame() -> bacteria_grid_df
dim(bacteria_grid_df)
n_sim = length(x = subsample_sbm_list)
```

```{r}
for(sim in 1:5){
  print(plotMyMatrix(log1p(as.matrix(subsample_list[[sim]]))))
}

```


```{r}
map(1:n_sim, \(x) data.frame(subsample = subsample_sbm_list[[x]]$memberships$grid_code,
                             grid_code = colnames(subsample_list[[1]]),
                             sim = x)
    ) %>%
  bind_rows() %>%
  pivot_wider(names_from = sim, names_glue = "simulation{sim}", values_from = "subsample") %>%
  column_to_rownames("grid_code") -> cluster_df
```

```{r}
n_echantillons <- nrow(cluster_df)
grid_cluster_mat <- matrix(0, nrow = n_echantillons, ncol = n_echantillons)

for (sim in 1:n_sim) {
  grid_cluster_mat <- grid_cluster_mat + outer(cluster_df[[sim]], cluster_df[[sim]], "==")
}

rownames(grid_cluster_mat) <- rownames(cluster_df)
colnames(grid_cluster_mat) <- rownames(cluster_df)
grid_cluster_mat %>%
  as.data.frame() %>%
  rownames_to_column("grid_code") %>%
  pivot_longer(-grid_code, names_to = "name", values_to = "value") %>%
  # Create two temporary columns where the names are ALWAYS sorted
  # pmin/pmax ensures that (Sample_A, Sample_B) and (Sample_B, Sample_A) 
  # both become (Sample_A, Sample_B)
  mutate(
    p1 = pmin(grid_code, name),
    p2 = pmax(grid_code, name)
  ) %>%
  # Distinct now recognizes these rows as identical
  distinct(p1, p2, .keep_all = TRUE) %>%
  # Drop temporary columns
  select(grid_code, name, value) %>%
  ggplot() +
    geom_histogram(aes(value)) +
    theme_minimal() +
    labs(x = "N time grids are in the same group (after 1000 sbm)")

plotMyMatrix(grid_cluster_mat, c(row = "grids", col = "grids"))

```

There are cleare group of sample that are appearing. 
It could be interesting to see what are the cluster found by an other sbm on this matrix

```{r}
estimateSimpleSBM(grid_cluster_mat,model = "poisson") -> sbm_copartnership
```


```{r}
plot(sbm_copartnership)
```

```{r}
diff_sequencingdepth <- as.matrix(dist(colSums(bacteria_grid_df)))
colnames(grid_cluster_mat) == colnames(diff_sequencingdepth)
rownames(grid_cluster_mat) == rownames(diff_sequencingdepth)

estimateSimpleSBM(grid_cluster_mat,model = "poisson", covariates = list(diff_sequencingdepth)) -> sbm_copartnership_depth_covar
```


```{r}
plot(sbm_copartnership_depth_covar)
plot(sbm_copartnership)
```


Sequencing depth could biais a lot clustering of subsamples. Thus I will try to put sequecing depth as covariate of the SMB

# Sequencing depth as covariate

First : test with one subsample (number one).

```{r}
if(file.exists("outputs/heavy_computation_save/smb_offset_test_on_subsample_one.Rdata")){
  
  load("outputs/heavy_computation_save/smb_offset_test_on_subsample_one.Rdata")  
  
}else{ 
    
  subsample_list[[1]] %>% 
    as.matrix() %>%
    log1p %>% 
    estimateBipartiteSBM(
      netMat       = ., 
      model        = "poisson", 
      dimLabels    = c(row = "ASV", col = "grid_code"),
      covariates   = list(library_size = covar_mat), # Liste nommée de matrices
      estimOptions = list(plot = FALSE, verbosity = 0)
    ) -> test
    save(test, file = "outputs/heavy_computation_save/smb_offset_test_on_subsample_one.Rdata")
  }
```

Problem : this is really computationally heave, around 2h for one SBM

```{r}
plot(test)
plot(subsample_sbm_list[[1]])


subsample_sbm_list[[1]]$memberships$grid_code
test$memberships$grid_code
str(test$covarEffect)
str(test$covarArray)

test$connectParam


B <- as.data.frame(
    table(
      
      colnames(bacteria_grid_df),
      subsample_sbm_list[[1]]$memberships$grid_code ,
      test$memberships$grid_code
      )
  )

colnames(B) =  c( "grid_code","sbm_bact", "sbm_bact_covariate","Freq")

data.frame(library_size = sort(colSums(bacteria_grid_df), decreasing = TRUE),
           col_library_size = viridis(43)) %>%
  rownames_to_column("grid_code")-> col_library_size_df


B %>%
  left_join(col_library_size_df, by = "grid_code") -> B




w   <- which(B$Freq != 0)
B <- B[w, ]

B$library_size <- factor(B$library_size, 
                         levels = sort(unique(B$library_size), decreasing = TRUE))
alluvial(B[,c(5,2,3)],
         freq = B$Freq,
         col = B$col_library_size,
         alpha = 0.7,
         cex = 1.5,
         cex.axis = 1.5,
         xw = 0.2)
```


```{r}
B %>%
  group_by(sbm_bact) %>%
  summarise(mean_library_size = mean(as.numeric(as.character(library_size)))) %>%
  arrange(mean_library_size)

B %>%
  group_by(sbm_bact_covariate) %>%
  summarise(mean_library_size = mean(as.numeric(as.character(library_size)))) %>%
  arrange(mean_library_size)
```

Using sequencing depth as covariate of the SBM doesn't correct the effect of library size. It lead, however, to a some change in the clustering.




