---
title: "Laws of diversity and variation — bacteria & fungi"
output: html_document
date: "2026-05-05"
---

This Rmd applies the framework of Grilli (2020) "Laws of diversity and variation
in microbial communities" (`jacopogrilli/lawsdiv`) to the prairiom2026
cross-sectional grid-level data, treating **bacteria (16S)** and **fungi (ITS2)**
as the two "environments" that replace the ten environments of the original paper.
We adapt only the cross-sectional part of `EstimatePars.R` (no longitudinal data
here, no host-specific stratification, no 10-environment correlation matrix).

# Load packages

```{r setup, message = FALSE, warning = FALSE}
library(tidyverse)     # data wrangling
library(ggfortify)     # lawsdiv dep
library(lamW)          # Lambert W (lawsdiv dep)
library(GGally)        # lawsdiv dep
library(ggcorrplot)    # lawsdiv dep
library(pracma)        # erfc, used by fun_erf / estimatemean_f
library(phyloseq)      # to read ps_bacteria.Rdata / ps_fungi.Rdata
library(scales)        # for fancy labellers used in mytheme
library(RColorBrewer)
```

# Define theme and core estimators inline

```{r lawsdiv-helpers}
## --- project plotting theme (matches scripts/3_plant_community_structure.R) ---
main_theme <- theme_minimal() +
  theme(line = element_blank(),
        axis.line = element_line(colour = "black"),
        panel.border = element_blank(),
        axis.ticks = element_line(colour = "black"),
        axis.text.x = element_text(colour = "black", size = 22),
        axis.text.y = element_text(colour = "black", size = 22),
        legend.title = element_text(colour = "black", size = 20, hjust = 0.5),
        legend.text = element_text(colour = "black", size = 18),
        axis.title = element_text(size = 28),
        strip.text = element_text(colour = "black", size = 15, face = "italic"))

## --- copied from scripts/lawsdiv/Code/EstimatePars.R ---------------------
## erfc-based equation whose root gives the lognormal mean estimate
fun_erf <- function(mu, c, m1, m2) {
  sigma <- sqrt(-c * m1 + m2 + c * mu - m1 * mu)
  x <- (c - mu) / sigma / sqrt(2)
  f <- (mu - m1) * erfc(x) + exp(-x^2) * sqrt(2 / pi) * sigma
  return(f)
}

## solve fun_erf == 0 for mu, given threshold c and the first two moments of
## log(f) above c
estimatemean_f <- function(c, m1, m2) {
  mumin <- (c * m1 - m2) / (c - m1)
  muest <- uniroot(fun_erf, c = c, m1 = m1, m2 = m2,
                   interval = c(-50, mumin - 0.001), tol = 0.0001)$root
  return(muest)
}

## reference parabola for the rescaled MAD plot (lognormal in log-space)
simple_par <- function(x) { -x^2 }

## copied from scripts/lawsdiv/Code/ggplot_mytheme.R — used by the AFD y-axis
fancy_scientific <- function(l) {
  l <- format(l, scientific = TRUE)
  l <- gsub("e\\+00", "", l)
  l <- gsub("^(.*)e", "'\\1'e", l)
  l <- gsub("'1'e", "10^", l)
  l <- gsub("e", "%*%10^", l)
  l <- gsub("\\+", "", l)
  parse(text = l)
}
```

# Load and aggregate data

We load both phyloseq objects, aggregate samples (quadrats) to grid level with
`merge_samples`, and drop the 2021 grids 14/15 — intentionally excluded
project-wide (see `CLAUDE.md`). After `merge_samples`, bacteria sample names
use `-` as separator while fungi sample names keep `_`, hence the two regex
variants.

```{r load-data}
load("outputs/clean_data_to_analise/ps_bacteria.Rdata")
load("outputs/clean_data_to_analise/ps_fungi.Rdata")

ps_bacteria_grid <- merge_samples(ps_bacteria, "grid") %>%
  prune_samples(!str_detect(sample_names(.), "21-CAM-1[45]"), .)

ps_fungi_grid <- merge_samples(ps_fungi, "grid_code") %>%
  prune_samples(!str_detect(sample_names(.), "21_CAM_1[45]"), .)

cat("Bacteria grid-level phyloseq:\n"); print(ps_bacteria_grid)
cat("\nFungi grid-level phyloseq:\n"); print(ps_fungi_grid)
```

```{r extract-otu-tables}
## Samples in columns, taxa in rows — same orientation as
## copartnership_bacteria_analysis.Rmd. After merge_samples, otu_table is
## samples-by-taxa, so we transpose.
bacteria_grid_df <- ps_bacteria_grid %>% otu_table() %>% t() %>% as.data.frame()
fungi_grid_df    <- ps_fungi_grid    %>% otu_table() %>% t() %>% as.data.frame()
cat("bacteria_grid_df: ", dim(bacteria_grid_df), "\n")
cat("fungi_grid_df:    ", dim(fungi_grid_df),    "\n")
```

# Build datatax (lawsdiv long format)

`EstimatePars.R` expects a long data frame with columns
`project_id, classification, otu_id, run_id, nreads, count`. Here:

- `project_id`     = `"prairiom2026"`
- `classification` = `"bacteria"` or `"fungi"` (the two environments)
- `otu_id`         = OTU/ASV row name
- `run_id`         = grid code (sample after merge)
- `count`          = integer reads for that OTU in that grid
- `nreads`         = column sum of counts across all OTUs in that grid

We compute `nreads` per `(classification, run_id)` from the wide matrix
**before** dropping zeros, then `pivot_longer` and join. We then filter
`count > 0` so only present OTUs occupy rows — this matches the effective
behaviour of `count > count_cutoff` with `count_cutoff = 0` in the original
script.

```{r build-datatax}
to_long <- function(wide_df, classification) {
  nreads_per_run <- tibble(
    run_id = colnames(wide_df),
    nreads = colSums(wide_df)
  )
  wide_df %>%
    rownames_to_column("otu_id") %>%
    pivot_longer(-otu_id, names_to = "run_id", values_to = "count") %>%
    mutate(
      project_id     = "prairiom2026",
      classification = classification,
      count          = as.integer(round(count))
    ) %>%
    left_join(nreads_per_run, by = "run_id") %>%
    select(project_id, classification, otu_id, run_id, nreads, count)
}

datatax <- bind_rows(
  to_long(bacteria_grid_df, "bacteria"),
  to_long(fungi_grid_df,    "fungi")
) %>%
  filter(count > 0)

cat("datatax dimensions: ", dim(datatax), "\n")
head(datatax)

datatax %>%
  group_by(classification) %>%
  summarise(n_runs = n_distinct(run_id),
            n_otus = n_distinct(otu_id),
            mean_nreads = mean(nreads),
            median_nreads = median(nreads)) %>%
  print()
```

# Filter and compute proj (per-OTU statistics)

We apply lawsdiv's mild cross-sectional cutoffs (`nreads > 10^1`) and compute
per-OTU statistics: occupancy `o` (fraction of grids where the OTU is present),
mean relative abundance `f` and variance `vf` of relative abundance across
grids. The block below is structurally identical to lines ~28-33 of
`EstimatePars.R`. We assign `sname = classification` directly (no `shortnames`
lookup — only two kingdoms here) and tag `scat` for downstream
classification-aware plots if needed.

```{r compute-proj}
nreads_cutoff <- 10^1
count_cutoff  <- 0

summarydata <- datatax %>%
  filter(nreads > nreads_cutoff) %>%
  group_by(project_id, classification) %>%
  summarise(n_of_runs = n_distinct(run_id),
            mean_nreads = mean(nreads), .groups = "drop") %>%
  filter(n_of_runs > 0) %>%
  mutate(idall = paste(project_id, classification))
print(summarydata)

datatax <- datatax %>%
  filter(paste(project_id, classification) %in%
         paste(summarydata$project_id, summarydata$classification))

proj <- datatax %>%
  filter(nreads > nreads_cutoff, count > count_cutoff) %>%
  group_by(project_id, classification, otu_id) %>%
  mutate(
    ## tf: mean relative abundance (count/nreads) over grids where the OTU IS present
    tf   = mean(count / nreads),
    ## o: raw count of grids where this OTU is present (fraction computed below)
    o    = n(),
    ## tvpf: shot-noise-corrected mean of p² over present grids.
    ##   Using (count²-count) instead of count² removes Poisson sampling noise:
    ##   E[k(k-1)/N²] ≈ p²  whereas  E[k²/N²] = p² + p/N (biased upward)
    tvpf = mean((count^2 - count) / nreads^2)
  ) %>%
  ungroup() %>%
  group_by(project_id, classification) %>%
  ## Convert o from raw count to fraction: n_present_grids / n_total_grids
  mutate(o = o / n_distinct(run_id)) %>%
  ## f  = E[p over ALL grids] = P(present) × E[p | present]  (absent grids contribute 0)
  ## vf = E[p² over ALL grids], same trick: multiply within-present mean by occupancy
  mutate(f = o * tf, vf = o * tvpf) %>%
  ## Complete the Var[p] = E[p²] - E[p]² identity → vf is now the population variance
  ## of relative abundance across ALL grids (zeros included, shot noise removed)
  mutate(vf = vf - f^2) %>%
  ungroup() %>%
  mutate(idall = paste(project_id, classification),
         sname = classification,
         scat  = classification) %>%
  select(-tf, -tvpf)

cat("proj dimensions: ", dim(proj), "\n")
head(proj)
summary(proj %>% select(o, f, vf))
```

# Gamma parameters (Abundance Fluctuation Distribution)

The Grilli framework predicts a Gamma AFD across samples, parameterised by
shape `beta = 1 / cv^2` (with `cv^2 = vf / f^2`) and scale `theta = f / beta`.

```{r gamma-pars}
gamma_pars <- proj %>%
  select(idall, sname, otu_id, o, f, vf) %>%
  ## cv = coefficient of variation of relative abundance across grids = sd/mean
  mutate(cv = sqrt(vf / f^2)) %>%
  distinct() %>%
  ## beta  = shape of the Gamma AFD = 1/cv²
  ##         high beta → OTU is stable across grids (low relative variability)
  ##         low beta  → OTU fluctuates strongly between grids
  ## theta = scale of the Gamma AFD = f/beta  (so that mean = beta*theta = f)
  mutate(beta = 1 / cv^2, theta = f / beta)

hist(log(gamma_pars$beta), breaks = 30)
cat("gamma_pars dimensions: ", dim(gamma_pars), "\n")
head(gamma_pars)

gamma_pars %>%
  group_by(sname) %>%
  summarise(n_otu = n(),
            n_with_var = sum(vf > 0, na.rm = TRUE),
            median_beta = median(beta[vf > 0], na.rm = TRUE)) %>%
  print()
```

# Taylor's law plot

Taylor's law states `Var(f) ~ <f>^2`. We plot it per kingdom (with the slope-2
reference line) and overlay binned means in the style of `Fig1.R`.

```{r taylor-plot, fig.width = 10, fig.height = 5}
nbins <- 15
## Build binned means to overlay on the Taylor plot (same binning trick as the AFD histogram)
## Goal: show the average trend through the cloud of per-OTU points
taylor_binnes <- gamma_pars %>%
  filter(f > 0, vf > 0) %>%
  group_by(idall, sname) %>%
  ## dlf = bin width in log(f) space; b = bin index for each OTU
  mutate(lf = log(f), dlf = (max(lf) - min(lf)) / nbins) %>%
  mutate(b = as.integer((lf - min(lf)) / dlf)) %>%
  ungroup() %>%
  ## Average f and vf within each bin → one representative point per bin per kingdom
  group_by(idall, sname, b) %>%
  summarise(vf = mean(vf), f = mean(f), .groups = "drop") %>%
  ## xx = random jitter used only to shuffle draw order (avoids one kingdom always on top)
  mutate(xx = runif(n())) %>%
  arrange(xx)

ptaylor_ind <- gamma_pars %>%
  filter(f > 0, vf > 0) %>%
  ggplot() + main_theme +
  aes(x = f, y = vf) +
  geom_abline(slope = 2, size = 1.2) +
  geom_point(size = 0.8, alpha = 0.4, color = "gray40") +
  geom_point(data = taylor_binnes,
             aes(x = f, y = vf, color = as.factor(sname),
                 shape = as.factor(sname)),
             size = 3, stroke = 1.2) +
  scale_x_log10("Average relative abundance") +
  scale_y_log10("Variance of relative abundance across grids") +
  theme(legend.position = "none") +
  facet_wrap(~ sname, scales = "free")
ptaylor_ind

ts <- format(Sys.time(), "%Y%m%d_%H%M%S")
# ggsave(filename = sprintf("outputs/figures/lawsdiv_taylor_%s.pdf", ts),
#        plot = ptaylor_ind, width = 10, height = 5)
```

# MAD parameters (Mean Abundance Distribution)

We now estimate the lognormal-tail parameters `(mu, sigma)` of the MAD on the
log of the mean abundance `f`, restricted to OTUs above a kingdom-specific
cutoff `c`. The original paper hand-tunes `c` per environment; here we use a
**principled heuristic**: for each kingdom, `c = quantile(log(f), 0.25)` —
the 25th-percentile of `log(f)`. This excludes the rarest OTUs whose mean
estimates are noise-dominated, without arbitrary tuning. Manual tuning is
likely needed if the parabola fit looks bad on either kingdom (see prose
note below).

```{r mad-cutoffs}
nbin <- 20

cutoffs <- gamma_pars %>%
  filter(f > 0) %>%
  group_by(sname) %>%
  summarise(c = as.numeric(quantile(log(f), 0.25)), .groups = "drop")
cutoffs

estimate_mean <- gamma_pars %>%
  filter(f > 0) %>%
  mutate(lf = log(f)) %>%
  ## Attach the per-kingdom cutoff c and keep only OTUs above it
  ## (rare OTUs below c have noisy mean estimates and would bias the fit)
  left_join(cutoffs, by = "sname") %>%
  ungroup() %>%
  filter(lf > c) %>%
  group_by(idall, sname) %>%
  summarise(
    c      = unique(c),
    ## m1 = first moment of log(f) above cutoff  = mean of log(f)
    m1     = mean(lf),
    ## m2 = second moment of log(f) above cutoff = mean of log(f)²
    ##      together m1 and m2 constrain the lognormal mu and sigma via fun_erf
    m2     = mean(lf^2),
    ## ns_obs = number of OTUs observed above the cutoff
    ns_obs = n_distinct(otu_id),
    ## nf = total relative abundance pooled across those OTUs
    nf     = sum(f),
    .groups = "drop"
  ) %>%
  rowwise() %>%
  mutate(
    ## mu    = lognormal mean of log(f): root of fun_erf given c, m1, m2
    ##         this is the peak of the full MAD (including OTUs below cutoff)
    mu    = estimatemean_f(c, m1, m2),
    ## sigma = lognormal sd of log(f), derived analytically once mu is known
    sigma = sqrt(-c * m1 + m2 + c * mu - m1 * mu)
  ) %>%
  ungroup() %>%
  as.data.frame() %>%
  ## stot = estimated TOTAL species richness (including OTUs too rare to observe above c)
  ##        = ns_obs / P(log(f) > c)  where P is the lognormal CDF tail
  ##        erfc((c-mu)/sigma/√2) / 2 = P(log(f) > c) for a N(mu, sigma²)
  mutate(stot = 2 * ns_obs / erfc((c - mu) / sigma / sqrt(2)))

estimate_mean
```

> **Note** — the cutoff heuristic above (`q25 of log(f)`) is a defensible
> default, not a final choice. The original paper tunes `c` by eye per
> environment (e.g. `c = -14.27` for `gut1`, `c = -18.8` for `sludge`). If
> the rescaled-MAD points below visibly deviate from the `-x^2` parabola in
> either kingdom, revisit `cutoffs` and pick a stricter quantile (e.g. 0.5
> or 0.75) or hand-set values per kingdom.

# Rescaled MAD plot (parabola)

We bin `log(f)` into `nbin = 20` bins per kingdom, compute the empirical
density `p`, then plot the lawsdiv rescaling: `(lf - mu) / sqrt(2 sigma^2)`
on x and `p / [0.5 * erfc((mu-c)/sqrt(2)/sigma)] * sqrt(2 pi)` on a log y.
A perfect lognormal collapses on the parabola `y = -x^2` (in log10 units).

```{r mad-plot, fig.width = 10, fig.height = 5}
## statp: histogram of log(f) per kingdom, used to plot the empirical MAD
## cutoff = -100 is a placeholder (kept in summarise to survive the group_by);
## the real cutoff comes from estimate_mean joined below
statp <- gamma_pars %>%
  filter(f > 0) %>%
  mutate(lf = log(f), cutoff = -100) %>%
  group_by(idall) %>%
  ## df = bin width in log(f) space (range / nbin)
  mutate(df = (max(lf) - min(lf)) / nbin) %>%
  ## b  = bin index: which of the nbin bins this OTU falls in
  mutate(b = as.integer((lf - min(lf)) / df)) %>%
  ungroup() %>%
  group_by(idall, b, df) %>%
  ## lf collapses to the bin mean; n = number of OTUs in this bin
  summarise(lf = mean(lf), cutoff = mean(cutoff), n = n(), .groups = "drop") %>%
  group_by(idall) %>%
  ## p = probability density = relative frequency / bin width
  ##     dividing by df converts probability (sums to 1) to density (integrates to 1)
  ##     so the histogram can be compared directly to the lognormal curve
  mutate(p = n / sum(n) / df) %>%
  ungroup()

p_mad <- statp %>%
  mutate(xx = runif(nrow(statp))) %>%
  filter(lf > cutoff) %>%
  left_join(estimate_mean, by = "idall") %>%
  arrange(xx, sname) %>%
  ggplot() + main_theme +
  ## Reference parabola: y = -x² is what a perfect lognormal gives in log10 units
  stat_function(fun = simple_par, color = "black", size = 1) +
  aes(
    ## x-axis: standardise log(f) to lognormal units → x=0 at the mean mu
    x = (lf - mu) / sqrt(2 * sigma^2),
    ## y-axis: rescale density so the truncated lognormal integrates to 1 above cutoff;
    ##   the erfc term accounts for the probability mass below the cutoff c
    ##   the +0.5*log(2π) converts from natural-log to log10 probability density
    y = 10^(log(p) - log(0.5 * erfc((mu - c) / sqrt(2) / sigma)) +
            0.5 * log(2 * pi)),
    color = as.factor(sname),
    shape = as.factor(sname)
  ) +
  geom_point(size = 3, stroke = 1) +
  ## Left vertical line: position of the cutoff c in rescaled units
  geom_vline(aes(xintercept = (c - mu) / sqrt(2 * sigma^2))) +
  ## Right vertical line: x=0, i.e. the estimated lognormal mean mu
  geom_vline(aes(xintercept = 0)) +
  scale_x_continuous("Rescaled log average abundance") +
  scale_y_log10("Fraction of species") +
  theme(legend.position = "none") +
  facet_wrap(~ sname)
p_mad

# ggsave(filename = sprintf("outputs/figures/lawsdiv_mad_%s.pdf", ts),
#        plot = p_mad, width = 10, height = 5)
```

# AFD plot (Gamma fluctuations across grids)

For OTUs present in nearly all grids (`o > 0.999`), we compute the per-OTU
standardised log relative abundance `(log(f_i) - mean) / sd` across grids,
bin it, and overlay the predicted Gamma curve `gammalog(x; k)` from `Fig1.R`
(with `k = 1.7`, the empirical mean shape from the paper). The lognormal
reference (`-x^2 - log10(2 pi) / 2 + log10(e)`) is plotted as a dashed line
for comparison.

```{r afd-plot, fig.width = 10, fig.height = 5}
## gammalog: log10 probability density of the standardised Gamma AFD (Grilli 2020 eq.1)
##   x = rescaled log relative abundance (standardised by sqrt(trigamma(k)))
##   k = shape parameter; k=1.7 is the paper's cross-environment empirical value
## lognlog: log10 probability density of a standard normal — the lognormal reference
##   a perfect lognormal gives a parabola -x² in log10 units (same as the MAD plot)
gammalog <- function(x, k = 1.7) {
  (k * trigamma(k) * x - exp(sqrt(trigamma(k)) * x + digamma(k))) -
    log(gamma(k)) + k * digamma(k) + log10(exp(1))
}
lognlog <- function(x) { -x^2 - log10(2 * pi) / 2 + log10(exp(1)) }

nbin_afd <- 20
rla_asv <- proj %>%
  ## Only OTUs present every grid so that each OTU contributes
  ## many data points to build its own within-OTU distribution
  filter(o > 0.999, nreads > 0, count > 0) %>%
  ## l      = log relative abundance for this (OTU, grid) observation
  ## f_inst = raw relative abundance for this observation (used for k below)
  mutate(l = log(count / nreads), f_inst = count / nreads) %>%
  group_by(otu_id, idall, sname) %>%
  ## ml = mean of log relative abundance across grids (per OTU)
  ## sl = sd   of log relative abundance across grids (per OTU)
  ## k  = empirical Gamma shape from moments: mean²/variance of relative abundance
  ##      (per-OTU estimate; the curve uses k=1.7 from Grilli, not this value)
  mutate(ml = mean(l), sl = sd(l),
         k  = mean(f_inst)^2 / var(f_inst)) %>%
  ungroup() %>%
  filter(!is.na(sl), sl > 0) %>%
  ## lf = standardised log relative abundance: (l - ml) / sl
  ##      centres each OTU at 0 and scales to unit variance
  ##      → all OTUs collapse onto the same x-axis regardless of their mean abundance
  mutate(lf = (l - ml) / sl)

dh_per_asv <- rla_asv %>%
  group_by(idall, otu_id, sname) %>%
  ## Manual histogram over the pooled standardized values:
  ## df = bin width, b = bin index (same logic as MAD histogram above)
  mutate(df = (max(lf) - min(lf)) / nbin_afd) %>%
  mutate(b  = as.integer((lf - min(lf)) / df)) %>%
  ungroup() %>%
  group_by(idall, b, df, sname,otu_id) %>%
  ## lf collapses to bin mean; n = observations per bin; k = mean shape in that bin
  summarise(lf = mean(lf), n = n(), k = mean(k, na.rm = TRUE),
            .groups = "drop") %>%
  group_by(idall) %>%
  ## p = probability density (same formula as MAD: relative frequency / bin width)
  mutate(p = n / sum(n) / df) %>%
  ungroup()
 
dh <- rla_asv %>%
  group_by(idall, sname) %>%
  ## Manual histogram over the pooled standardized values:
  ## df = bin width, b = bin index (same logic as MAD histogram above)
  mutate(df = (max(lf) - min(lf)) / nbin_afd) %>%
  mutate(b  = as.integer((lf - min(lf)) / df)) %>%
  ungroup() %>%
  group_by(idall, b, df, sname) %>%
  ## lf collapses to bin mean; n = observations per bin; k = mean shape in that bin
  summarise(lf = mean(lf), n = n(), k = mean(k, na.rm = TRUE),
            .groups = "drop") %>%
  group_by(idall) %>%
  ## p = probability density (same formula as MAD: relative frequency / bin width)
  mutate(p = n / sum(n) / df) %>%
  ungroup()

cat("dh dimensions (binned standardised AFD points): ", dim(dh), "\n")
```


```{r afd-plot, fig.width = 10, fig.height = 5}
dh %>%
  arrange(sname) %>%
  ggplot(   
    aes(x = lf / sqrt(2))) +
   aes(

    ## y = p on a log10 scale to overlay the gammalog / lognlog curves directly
    y = p,
    color = as.factor(sname),
    shape = as.factor(sname)
  ) +
  stat_function(fun = function(x) 10^lognlog(x), color = "black", size = 1, linetype = "dashed") +
  stat_function(fun = function(x) 10^gammalog(x), color = "black", size = 1) +
 
  geom_point(size = 3, stroke = 1.3) +
  scale_x_continuous("Rescaled log relative abundance") +
  scale_y_log10("Probability density", labels = fancy_scientific) +
  theme(legend.position = "none") +
  facet_wrap(~ sname) +
  main_theme



dh_per_asv %>%
  ggplot() +
    aes(
    ## x = lf/√2: extra rescaling to match the Grilli paper's x-axis convention
    ##     (their gammalog uses sqrt(trigamma(k)) as the scale, which ≈ √2 at k=1.7)
    x = lf / sqrt(2),
    ## y = p on a log10 scale to overlay the gammalog / lognlog curves directly
    y = p,
    color = as.factor(sname),
    shape = as.factor(sname),
    group = otu_id
  ) +
 stat_function(fun = function(x) 10^lognlog(x), color = "black", linetype = "dashed") +
  stat_function(fun = function(x) 10^gammalog(x), color = "black") +

  geom_line(size = 1, stroke = 1.3, alpha =0.3) +
  scale_x_continuous("Rescaled log relative abundance") +
  scale_y_log10("Probability density", labels = fancy_scientific) +
  theme(legend.position = "none") +
  facet_wrap(~ sname) +
  main_theme 
```

If `dh` is empty (no OTU present in essentially every grid), the AFD plot
above will not render meaningful points — likely with only ~40 grids per
kingdom and the strict `o > 0.999` filter. In that case relax the filter
(e.g. `o > 0.5`) or reduce `nbin_afd`; we keep the strict filter here to
match the original paper's figure.

# Mean parameters table

`mean_pars` collects per-kingdom MAD parameters together with the mean Gamma
shape `mbeta` (averaged over OTUs with `f > 1e-5` and `vf > 0`). This is the
final summary table consumed downstream.

```{r mean-pars}
mean_pars <- estimate_mean %>%
  ## MAD parameters per kingdom: c (cutoff), mu and sigma (lognormal), stot (total richness)
  select(idall, sname, nf, c, mu, sigma, stot) %>%
  left_join(
    gamma_pars %>%
      filter(f > 1e-5, vf > 0) %>%
      group_by(sname) %>%
      ## mbeta = mean Gamma shape across OTUs = average stability of abundance fluctuations
      ##         higher mbeta → community tends toward stable, less variable OTUs
      summarise(mbeta = mean(beta), .groups = "drop"),
    by = "sname"
  )

mean_pars
```

# Simulate community

```{r}
gamma_pars %>%
  mutate(intercept_tay = log(vf) - 2*log(f),
         beta2 = 1/exp(intercept_tay)) 
```
Same beta tahn the already computed one

## Bacteria

```{r}
gamma_pars %>%
  filter(bacteria) %>%
  
```

