---
title: "Untitled"
output: html_document
date: "2026-05-15"
---
# Load packages

```{r setup, message = FALSE, warning = FALSE}
library(tidyverse)     # data wrangling
library(ggfortify)     # lawsdiv dep
library(lamW)          # Lambert W (lawsdiv dep)
library(GGally)        # lawsdiv dep
library(ggcorrplot)    # lawsdiv dep
library(pracma)        # erfc, used by fun_erf / estimatemean_f
library(phyloseq)      # to read ps_bacteria.Rdata / ps_fungi.Rdata
library(scales)        # for fancy labellers used in mytheme
library(RColorBrewer)
library(broom)         # to reforam lm of taylor's law
```

# Define theme and core estimators inline

```{r lawsdiv-helpers}
## --- project plotting theme (matches scripts/3_plant_community_structure.R) ---
main_theme <- theme_minimal() +
  theme(line = element_blank(),
        axis.line = element_line(colour = "black"),
        panel.border = element_blank(),
        axis.ticks = element_line(colour = "black"),
        axis.text.x = element_text(colour = "black", size = 22),
        axis.text.y = element_text(colour = "black", size = 22),
        legend.title = element_text(colour = "black", size = 20, hjust = 0.5),
        legend.text = element_text(colour = "black", size = 18),
        axis.title = element_text(size = 28),
        strip.text = element_text(colour = "black", size = 15, face = "italic"))
```

# Load and aggregate data

```{r load-data}
load("outputs/clean_data_to_analise/ps_bacteria.Rdata")
load("outputs/clean_data_to_analise/ps_fungi.Rdata")

## merge_samples() default fun = sum -> grid-level reads are SUMMED across
## quadrats, not averaged. Matches lawsdiv_bact_fungi_analysis.Rmd; switching
## to fun = mean would require re-running that script too.
ps_bacteria_grid <- merge_samples(ps_bacteria, "grid") %>%
  prune_samples(!str_detect(sample_names(.), "21-CAM-1[45]"), .)

ps_fungi_grid <- merge_samples(ps_fungi, "grid_code") %>%
  prune_samples(!str_detect(sample_names(.), "21_CAM_1[45]"), .)

cat("Bacteria grid-level phyloseq:\n"); print(ps_bacteria_grid)
cat("\nFungi grid-level phyloseq:\n"); print(ps_fungi_grid)
```

```{r extract-otu-tables}
bacteria_grid_df <- ps_bacteria_grid %>% otu_table() %>% t() %>% as.data.frame()
fungi_grid_df    <- ps_fungi_grid    %>% otu_table() %>% t() %>% as.data.frame()
cat("bacteria_grid_df: ", dim(bacteria_grid_df), "\n")
cat("fungi_grid_df:    ", dim(fungi_grid_df),    "\n")
```

# Build datatax (long format)

Long data frame with columns:

- `project_id`     = `"prairiom2026"`
- `classification` = `"bacteria"` or `"fungi"` (the two environments)
- `otu_id`         = OTU/ASV row name
- `grid_code`      = grid code (sample after `merge_samples`)
- `nreads`         = integer reads for that OTU in that grid
- `tot_reads`      = column sum of reads across all OTUs in that grid

`tot_reads` is computed from the wide matrix **before** dropping zeros (so
absent OTUs still contribute their zero to the column total — i.e. don't
inflate the total), then joined back. We then filter `nreads > 0` so only
present OTUs occupy rows.

```{r build-datatax}
## Pivot wide OTU x grid matrix to long, attach grid-level read totals via a
## pre-computed join (avoids leaving the output silently grouped by grid_code).
to_long <- function(wide_df, classification) {

  tot_reads_per_grid <- tibble(
    grid_code = colnames(wide_df),
    tot_reads = colSums(wide_df)
  )

  wide_df %>%
    rownames_to_column("otu_id") %>%
    pivot_longer(-otu_id, names_to = "grid_code", values_to = "nreads") %>%
    left_join(tot_reads_per_grid, by = "grid_code") %>%
    mutate(project_id     = "prairiom2026",
           classification = classification)
}
```

# Compute moments

Working on presence only to compute mean and variance of relative abundance
conditional on presence:

$$\mathbb{E}[X \mid \text{present}]$$

```{r apply-to-long}
datatax <- bind_rows(
  to_long(bacteria_grid_df, "bacteria"),
  to_long(fungi_grid_df,    "fungi")
) %>%
  filter(nreads > 0)

cat("datatax dimensions: ", dim(datatax), "\n")
head(datatax)

datatax %>%
  group_by(classification) %>%
  summarise(n_runs = n_distinct(grid_code),
            n_otus = n_distinct(otu_id),
            mean_nreads = mean(nreads),
            median_nreads = median(nreads)) %>%
  print()
```

$$\mathbb{E}[X] = P(\text{present}) \cdot \mathbb{E}[X \mid \text{present}] + P(\text{absent}) \cdot \mathbb{E}[X \mid \text{absent}]$$
$$ \mathbb{E}[X \mid \text{absent}] = 0$$
$$\mathbb{E}[X] = P(\text{present}) \cdot \mathbb{E}[X \mid \text{present}]$$

```{r moment-df}
reads_tot_cutoff <- 10^1

## Per-OTU statistics computed on grids where the OTU is present (datatax was
## already filtered to nreads > 0), then re-weighted by occupancy o to recover
## unconditional moments via the law of total expectation.
##   cond_mean_rel_ab = E[p | present]
##   cond_s_moment    = E[p^2 | present] with bias correction (n^2 -> n*(n-1)),
##                      which removes the Poisson shot-noise term E[p/N]
##   mean_rel_ab      = o * E[p | present]              = E[p]
##   s_moment         = o * E[p^2 | present]            = E[p^2]
##   var_rel_ab       = E[p^2] - E[p]^2                 = Var[p]
moment_df <- datatax %>%
  filter(tot_reads > reads_tot_cutoff) %>%
  group_by(classification) %>%
  mutate(n_runs = n_distinct(grid_code)) %>%   # total grids per kingdom (constant within group)
  group_by(classification, otu_id) %>%
  summarise(n_runs           = first(n_runs),
            cond_mean_rel_ab = mean(nreads / tot_reads),
            o                = n() / first(n_runs),
            cond_s_moment    = mean((nreads^2 - nreads) / tot_reads^2),
            .groups = "drop") %>%
  mutate(mean_rel_ab = cond_mean_rel_ab * o,
         s_moment    = cond_s_moment    * o,
         var_rel_ab  = s_moment - mean_rel_ab^2)
```

# Taylor's law

Taylor's law predicts $\mathrm{Var}[p] \propto \mathbb{E}[p]^{2}$ (slope 2 in
log–log). The black reference line has slope 2; the blue line is the fitted
regression per kingdom. Slope close to 2 → Taylor's law holds and a constant
β makes sense; large deviation → the constant-β framework is shaky for that
kingdom.

```{r taylor-plot}
moment_df %>%
  filter(mean_rel_ab > 0, var_rel_ab > 0) %>%
  ggplot(aes(mean_rel_ab, var_rel_ab)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE, colour = "steelblue") +
  geom_abline(slope = 2, linetype = "dashed") +
  facet_wrap(~classification) +
  scale_x_log10() +
  scale_y_log10() +
  main_theme
```

```{r taylor-fit}
## Formal slope test per kingdom: log(Var[p]) ~ log(E[p]).
## Expected slope = 2 under Taylor's law.
taylor_fits <- moment_df %>%
  filter(mean_rel_ab > 0, var_rel_ab > 0) %>%
  nest(data = -classification) %>%
  mutate(fit    = map(data, ~ lm(log(var_rel_ab) ~ log(mean_rel_ab), data = .x)),
         tidy   = map(fit, broom::tidy, conf.int = TRUE)) %>% # turn fit into a tidy table
  select(classification, tidy) %>%
  unnest(tidy)

taylor_fits   # slope = row with term == "log(mean_rel_ab)"
```
# Gamma

Per-OTU Gamma shape $\hat\beta_i = \bar{x}_i^2 / \sigma_{x_i}^2$ and
scale $\hat\theta_i = \bar{x}_i / \hat\beta_i$. Under Taylor's law the
$\hat\beta_i$ collapse to a single constant β; we take the median as a
robust point estimate. Low-occupancy OTUs have noisy variance estimates, so
we also compute the median restricted to occupancy `o > 0.5` (Grilli Fig 1c
style) — useful as a sensitivity check.

```{r gamma-pars}
gamma_pars <- moment_df %>%
  mutate(beta  = mean_rel_ab^2 / var_rel_ab,
         theta = mean_rel_ab / beta)

## Log-beta distribution split by occupancy. Unimodal & tight => constant-β
## holds; bimodal / heavy-tailed => low-occupancy OTUs are distorting things.
gamma_pars %>%
  filter(beta > 0, is.finite(beta)) %>%
  mutate(occupancy = if_else(o > 0.5, "high (o > 0.5)", "low (o <= 0.5)")) %>%
  ggplot(aes(log(beta), fill = occupancy)) +
  geom_histogram(bins = 40, position = "identity", alpha = 0.5) +
  facet_wrap(~classification, scales = "free_y") +
  main_theme
```

```{r beta-med}
## Both options computed; choose one at simulation time.
beta_med <- gamma_pars %>%
  group_by(classification) %>%
  summarise(n_otu               = n(),
            n_with_var          = sum(mean_rel_ab > 0 & o > 0.5, na.rm = TRUE),
            median_beta_all     = median(beta[mean_rel_ab > 0], na.rm = TRUE),
            median_beta_highocc = median(beta[mean_rel_ab > 0 & o > 0.5], na.rm = TRUE),
            .groups = "drop")

beta_med
```

# MAD

Lognormal MAD: the mean relative abundances $\bar{x}_i$ are assumed
$\log\bar{x}_i \sim \mathcal{N}(\mu, \sigma^2)$ above a detection cutoff $c$.
We fit $(\mu, \sigma)$ by truncated-lognormal MLE and then estimate the total
number of species $s_\text{tot}$ in two ways (compared below).

```{r mad-hist}
## Diagnostic — log(mean_rel_ab) should be roughly bell-shaped above the cutoff
## for the truncated-lognormal MLE to be meaningful.
gamma_pars %>%
  filter(mean_rel_ab > 0) %>%
  ggplot(aes(log(mean_rel_ab))) +
  geom_histogram(bins = 40) +
  facet_wrap(~classification, scales = "free") +
  main_theme
```

```{r fun-erf}
## Truncated-lognormal MLE (Grilli SI Eqs. 50-52): two equations in (mu, sigma)
## reduced to 1D by algebraically eliminating sigma from the second equation.
## fun_erf returns the residual of the first MLE equation; uniroot finds mu.
fun_erf <- function(mu, c, m1, m2){

  sigma <- sqrt(m2 - m1*mu - c * m1 + mu * c)

  z <- (c - mu) / (sqrt(2) * sigma)
  correction <- sqrt(2 / pi) * sigma * exp(-z^2) / erfc(z)

  res1 <- m1 - (correction + mu)
  return(res1)
}

estimatemean_mu <- function(m1, m2, c){
  mumin <- (c * m1 - m2) / (c - m1)
  muest <- uniroot(fun_erf, c = c, m1 = m1, m2 = m2,
                   interval = c(-50, mumin - 0.001), tol = 0.0001)$root
}
```

```{r mad-fit}
## Cutoff c = 10th percentile of log(mean_rel_ab) per kingdom — drops the
## sparsely-observed bottom tail where the empirical mean is unreliable.
cutoffs <- gamma_pars %>%
  group_by(classification) %>%
  summarise(c = as.numeric(quantile(log(mean_rel_ab), 0.1)), .groups = "drop")

## Both s_tot definitions are computed; pick one at simulation time.
##   s_tot_comp (SI Eq. 55) — compositional: <xi>·s_tot = 1 => s_tot = 1/<xi>.
##                            Sensitive to sigma; often unstable.
##   s_tot_erfc (SI Eq. 53) — inverts the truncated-lognormal CDF using the
##                            number of species observed above c. Preferred.
mad_pars <- gamma_pars %>%
  left_join(cutoffs, by = "classification") %>%
  mutate(log_mean_rel_ab = log(mean_rel_ab)) %>%
  filter(log_mean_rel_ab > c) %>%
  group_by(classification, c) %>%
  summarise(m1    = mean(log_mean_rel_ab),
            m2    = mean(log_mean_rel_ab^2),
            S_obs = n_distinct(otu_id),
            .groups = "drop") %>%
  rowwise() %>%   # uniroot is scalar; force per-row evaluation
  mutate(mu         = estimatemean_mu(m1, m2, c),
         sigma      = sqrt(m2 - m1*mu - c*m1 + mu*c),
         s_tot_comp = 1 / exp(mu + sigma^2/2),
         s_tot_erfc = 2 * S_obs / erfc((c - mu) / (sqrt(2) * sigma))) %>%
  ungroup()

mad_pars   # compare s_tot_comp vs s_tot_erfc per kingdom
```

```{r s-tot-observed}
## Total observed OTUs per kingdom (no cutoff) — reference value to compare
## against the MAD-extrapolated s_tot_comp / s_tot_erfc.
S_tot <- gamma_pars %>%
  group_by(classification) %>%
  summarise(S_observed = n_distinct(otu_id), .groups = "drop")

S_tot
```
# Simulate communities

Each kingdom is simulated independently with its own fitted parameters and its
own observed grid-level read depths. The two `# ----` toggles at the top of
the `map()` body switch between (i) the two `s_tot` definitions and (ii) the
two β medians.

Result lives in `simulations`, a named list:
- `simulations$bacteria$xi`                — drawn species pool (relative abund.)
- `simulations$bacteria$rel_abundance_sim` — species × grid matrix of realised
                                              relative abundances (gamma AFD)
- `simulations$bacteria$abundance_sim`     — species × grid integer reads
                                              (multinomial)
- `simulations$bacteria$abundance_sim_final` — same, with undetected species
                                                dropped
(same structure under `simulations$fungi`).

## Grid-level read depths (used to size the multinomial step)
```{r tot-reads-df}
## tot_reads is already a column in datatax 
tot_reads_df <- datatax %>%
  distinct(classification, grid_code, tot_reads)
```

## Run the simulation for both kingdoms
```{r simulate}
set.seed(2026)

simulations <- c("bacteria", "fungi") %>%
  set_names() %>%
  map(function(k) {

    pars  <- mad_pars     %>% filter(classification == k)
    bpars <- beta_med     %>% filter(classification == k)
    grids <- tot_reads_df %>% filter(classification == k)

    ## ---- pick s_tot definition (uncomment one) ----
    s_pool <- pars$s_tot_comp
    # s_pool <- pars$s_tot_erfc

    ## ---- pick beta (uncomment one) ----
    beta_use <- bpars$median_beta_all
    # beta_use <- bpars$median_beta_highocc

    ## Step A — species pool from lognormal MAD, renormalised so sum(xi) = 1.
    xi <- exp(rnorm(round(s_pool), mean = pars$mu, sd = pars$sigma))
    xi <- xi / sum(xi)

    ## Step B — gamma realised relative abundances per grid
    ##          (shape = beta, mean = xi), then column-normalised to sum to 1.
    rel_abundance_sim <- replicate(
      nrow(grids),
      rgamma(length(xi), shape = beta_use, scale = xi / beta_use)
    ) %>% prop.table(margin = 2)

    ## Step C — multinomial reads using each grid's real read depth.
    abundance_sim <- map2(
      grids$tot_reads, asplit(rel_abundance_sim, 2),
      ~ rmultinom(1, size = .x, prob = .y)
    ) %>% do.call(cbind, .)
    colnames(abundance_sim) <- grids$grid_code

    ## Step D — drop species never detected (mirrors what real sampling produces).
    abundance_sim_final <- abundance_sim[rowSums(abundance_sim) > 0, ]

    list(xi                  = xi,
         rel_abundance_sim   = rel_abundance_sim,
         abundance_sim       = abundance_sim,
         abundance_sim_final = abundance_sim_final)
  })

```

## Sanity checks

```{r sim-shape}
## Simulated vs observed matrix shapes. Column count must match (grids);
## row count differs because the species pool is the extrapolated s_tot
## and undetected species are dropped at the end.
list(bacteria_obs = dim(bacteria_grid_df),
     bacteria_sim = dim(simulations$bacteria$abundance_sim_final),
     fungi_obs    = dim(fungi_grid_df),
     fungi_sim    = dim(simulations$fungi$abundance_sim_final))
```

```{r sim-depths}
## rmultinom preserves the size argument exactly, so simulated column sums
## should equal the observed read depths.
all.equal(unname(colSums(simulations$bacteria$abundance_sim)),
          tot_reads_df %>% filter(classification == "bacteria") %>% pull(tot_reads))

all.equal(unname(colSums(simulations$fungi$abundance_sim)),
          tot_reads_df %>% filter(classification == "fungi") %>% pull(tot_reads))
```

```{r}
library(sbm)

plotMyMatrix(t(log1p(simulations$fungi$abundance_sim_final)))
```

