---
title: "Bipartite SBM per Kingdom"
author: "Mathis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  cache   = FALSE
)
```

# Packages and helpers

```{r packages}
library(tidyverse)
library(phyloseq)
library(sbm)
library(iNEXT.3D)
library(patchwork)
library(alluvial)
library(PLNmodels)


source("scripts/plot_multipartite_sbm.R")
source("scripts/functions_modified.R")
```

---

# Data loading

All four kingdoms are aggregated to the **grid level** (sampling unit: `YY_CAM_GG`).
Grids 14 and 15 from 2021 are excluded throughout.

## Bacteria

```{r load-bacteria}
load("outputs/clean_data_to_analise/ps_bacteria.Rdata")

ps_bacteria_grid <- merge_samples(ps_bacteria, "grid") %>%
  prune_samples(!str_detect(sample_names(.), "21-CAM-1[45]"), .)

# OTU table: rows = OTUs, columns = grids
ps_bacteria_grid %>%
  otu_table() %>%
  t() %>%
  as.data.frame() %>%
  rename_with(~str_replace_all(., "-", "_")) -> bacteria_grid_matrix

# The column order of bacteria defines the reference sort for all other kingdoms
reference_sample_sort <- colnames(bacteria_grid_matrix)

cat("Bacteria grid matrix:", nrow(bacteria_grid_matrix), "OTUs x", ncol(bacteria_grid_matrix), "grids\n")
```

## Fungi

```{r load-fungi}
load("outputs/clean_data_to_analise/ps_fungi.Rdata")

ps_fungi_grid <- merge_samples(ps_fungi, "grid_code") %>%
  prune_samples(!str_detect(sample_names(.), "21_CAM_1[45]"), .)

# Align columns to bacteria sort order
ps_fungi_grid %>%
  otu_table() %>%
  t() %>%
  as.data.frame() -> fungi_grid_matrix

fungi_grid_matrix <- fungi_grid_matrix[, reference_sample_sort]

cat("Fungi grid matrix:", nrow(fungi_grid_matrix), "OTUs x", ncol(fungi_grid_matrix), "grids\n")
```

## Viruses

Virus data are at quadrat level; they are summed to grid level here.

```{r load-virus}
read.table("outputs/clean_data_to_analise/OTU_virus_CAM.txt", header = TRUE) %>%
  mutate(grid_code = str_sub(Host_code, 1, 9)) %>%
  filter(!str_detect(grid_code, "21_CAM_1[45]")) %>%
  select(-Host_code) %>%
  group_by(grid_code) %>%
  summarise(across(where(is.numeric), sum)) %>%
  column_to_rownames("grid_code") %>%
  as.matrix() -> virus_grid_matrix

# Align rows to reference sort
virus_grid_matrix <- virus_grid_matrix[reference_sample_sort, ]

cat("Virus grid matrix:", nrow(virus_grid_matrix), "grids x", ncol(virus_grid_matrix), "OTUs\n")
```

## Plants

Plant abundances are already aggregated at grid level (quadrat cover × 2, summed).

```{r load-plants}
read.table("outputs/clean_data_to_analise/abund_plant_grid.txt", header = TRUE) %>%
  column_to_rownames("Grid_code") %>%
  as.matrix() -> plant_grid_matrix

plant_grid_matrix <- plant_grid_matrix[reference_sample_sort, ]

cat("Plant grid matrix:", nrow(plant_grid_matrix), "grids x", ncol(plant_grid_matrix), "OTUs\n")
```

## Metadata

```{r load-metadata}
read.table("outputs/clean_data_to_analise/Metadata_grid_CAM.txt", header = TRUE) -> metadata_grid
cat("Metadata:", nrow(metadata_grid), "grids,", ncol(metadata_grid), "variables\n")
```

---

# SBM at grid level

## Read count truncation

High sequencing depth inflates the matrix and slows SBM fitting.
Reads are floored at 1/10 of their original value; zero-count entries are then
dropped to keep the matrix sparse.

```{r truncate-grid}
truncate_matrix <- function(otu_mat, divisor = 10) {
  otu_mat %>%
    rownames_to_column("OTU") %>%
    pivot_longer(-OTU, names_to = "sample", values_to = "reads") %>%
    mutate(reads = floor(reads / divisor)) %>%
    filter(reads > 0) %>%
    pivot_wider(names_from = "sample", values_from = "reads", values_fill = 0) %>%
    column_to_rownames("OTU") %>%
    # pivot_wider reorders columns based on first appearance after filtering;
    # restore the original column order explicitly
    .[, colnames(otu_mat)]
}

bacteria_grid_truncated <- truncate_matrix(bacteria_grid_matrix)
fungi_grid_truncated    <- truncate_matrix(fungi_grid_matrix)

cat("After truncation — Bacteria:", nrow(bacteria_grid_truncated), "OTUs\n")
cat("After truncation — Fungi   :", nrow(fungi_grid_truncated),    "OTUs\n")
```

## Coverage-based rarefaction

### Minimum observed coverage per kingdom

All samples within a kingdom are rarefied to the same coverage level — the
minimum observed sample coverage — so that richness comparisons are not biased
by sequencing depth.

```{r coverage-bacteria}
iNext_bact <- bacteria_grid_truncated %>%
  iNEXT3D(q = 0, datatype = "abundance", nboot = 0)

min_coverage_bact <- iNext_bact$TDiNextEst$coverage_based %>%
  filter(Order.q == 0, Method == "Observed") %>%
  slice_min(SC, n = 1) %>%
  pull(SC)

required_sizes_bact <- bacteria_grid_truncated %>%
  estimate3D(level = min_coverage_bact, q = 0, nboot = 0) %>%
  select(Assemblage, m)

cat("Minimum sample coverage (bacteria):", round(min_coverage_bact, 3), "\n")
```

```{r coverage-fungi}
iNext_fungi <- fungi_grid_truncated %>%
  iNEXT3D(q = 0, datatype = "abundance", nboot = 0)

min_coverage_fungi <- iNext_fungi$TDiNextEst$coverage_based %>%
  filter(Order.q == 0, Method == "Observed") %>%
  slice_min(SC, n = 1) %>%
  pull(SC)

required_sizes_fungi <- fungi_grid_truncated %>%
  estimate3D(level = min_coverage_fungi, q = 0, nboot = 0) %>%
  select(Assemblage, m)

cat("Minimum sample coverage (fungi):", round(min_coverage_fungi, 3), "\n")
```

Comparison of target rarefaction sizes between bacteria and fungi:

```{r coverage-comparison-plot, fig.width=5, fig.height=4}
required_sizes_fungi %>%
  left_join(required_sizes_bact, by = "Assemblage", suffix = c("_fungi", "_bact")) %>%
  ggplot(aes(m_fungi, m_bact)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Target size (fungi)", y = "Target size (bacteria)") +
  theme_bw()
```

### Rarefaction function

The function samples `target_size` reads without replacement from each sample,
repeating the process `n_sim` times. The per-OTU totals across simulations are
then accumulated into a single matrix (effectively averaging over stochasticity).

```{r rarefy-function}
coverage_rarefy_matrix <- function(otu_matrix, required_sizes) {
  sapply(colnames(otu_matrix), function(sample_name) {
    sample_vector <- otu_matrix[, sample_name]
    target_size <- required_sizes %>%
      filter(Assemblage == sample_name) %>%
      pull(m)

    # Pool all reads as individual OTU tags, then draw without replacement
    otu_pool    <- rep(rownames(otu_matrix), times = sample_vector)
    sampled     <- sample(otu_pool, size = target_size, replace = FALSE)
    rarefied    <- table(factor(sampled, levels = rownames(otu_matrix)))
    as.numeric(rarefied)
  }) %>%
    `rownames<-`(rownames(otu_matrix)) %>%
    as.data.frame() %>%
    filter(rowSums(across(where(is.numeric))) != 0) %>%
    as.matrix()
}
```

### Simulation (n = 100)

Each simulation draws a new rarefied matrix; the 100 matrices are stacked and
summed per (OTU, sample) cell. OTUs present in fewer than 4 reads on average
(i.e. total < 400 across simulations) are discarded as likely noise.

```{r rarefy-grid, eval=FALSE}
n_sim  <- 100
set.seed(2365)

accumulate_simulations <- function(sim_list, n_sim) {
  sim_list %>%
    map(~.x %>%
          as.data.frame() %>%
          rownames_to_column("OTU") %>%
          pivot_longer(-OTU, names_to = "sample", values_to = "count")) %>%
    reduce(full_join, by = c("OTU", "sample")) %>%
    mutate(total = rowSums(across(starts_with("count")), na.rm = TRUE)) %>%
    select(OTU, sample, total) %>%
    pivot_wider(names_from = "sample", values_from = "total") %>%
    column_to_rownames("OTU") %>%
    as.matrix()
}

# Bacteria
map(1:n_sim, ~coverage_rarefy_matrix(bacteria_grid_truncated,
                                     required_sizes_bact)) -> bact_sim_list
bact_grid_cumul <- accumulate_simulations(bact_sim_list, n_sim)
# Keep OTUs with mean abundance > 3 (total > 3 * n_sim)
bact_grid_rarefied <- bact_grid_cumul[rowSums(round(bact_grid_cumul / n_sim)) > 3, ]

# Fungi
map(1:n_sim, ~coverage_rarefy_matrix(fungi_grid_truncated,
                                     required_sizes_fungi)) -> fungi_sim_list
fungi_grid_cumul <- accumulate_simulations(fungi_sim_list, n_sim)
fungi_grid_rarefied <- fungi_grid_cumul[rowSums(round(fungi_grid_cumul / n_sim)) > 3, ]

cat("Retained OTUs — Bacteria:", nrow(bact_grid_rarefied),
    "| Fungi:", nrow(fungi_grid_rarefied), "\n")

# Sanity check: grid order must match across all kingdoms
stopifnot(
  all(colnames(bact_grid_rarefied) == colnames(fungi_grid_rarefied)),
  all(colnames(fungi_grid_rarefied) == rownames(virus_grid_matrix)),
  all(rownames(virus_grid_matrix)   == rownames(plant_grid_matrix))
)
```

## SBM fitting

Bipartite SBMs relate grids (row nodes) to OTUs (column nodes). The emission
model is chosen per kingdom:

- **Bacteria / Fungi**: Gaussian on log1p-transformed cumulated rarefied counts.
- **Virus**: Poisson on raw counts (low read depth, integers well-behaved).
- **Plants**: Gaussian on log1p-transformed abundance.

Heavy computation blocks below are set to `eval = FALSE`; cached objects are
loaded immediately after.

### Bacteria

```{r sbm-bacteria, eval=FALSE}
if(!file.exist("outputs/heavy_computation_save/sbm_Bact_gaussian.Rdata")){
  sbm_Bact_gaussian <- estimateBipartiteSBM(
  log1p(t(bact_grid_rarefied)),
  model     = "gaussian",
  dimLabels = c("grid", "bact")
)
save(sbm_Bact_gaussian, file = "outputs/heavy_computation_save/sbm_Bact_gaussian.Rdata")
}
```

```{r load-sbm-bacteria}
load("outputs/heavy_computation_save/sbm_Bact_gaussian.Rdata")
```

### Fungi

```{r sbm-fungi, eval=FALSE}
if(!file.exist("outputs/heavy_computation_save/sbm_Fungi_gaussian.Rdata")){
sbm_Fungi_gaussian <- estimateBipartiteSBM(
  log1p(t(fungi_grid_rarefied)),
  model     = "gaussian",
  dimLabels = c("grid", "fungi")
)
save(sbm_Fungi_gaussian, file = "outputs/heavy_computation_save/sbm_Fungi_gaussian.Rdata")
}
```

```{r load-sbm-fungi}
load("outputs/heavy_computation_save/sbm_Fungi_gaussian.Rdata")
```

### Viruses

```{r sbm-virus}
sbm_Virus <- estimateBipartiteSBM(
  virus_grid_matrix,
  model     = "poisson",
  dimLabels = c("grid", "virus")
)
```

### Plants

```{r sbm-plants}
sbm_Plant <- estimateBipartiteSBM(
  log1p(plant_grid_matrix),
  model     = "gaussian",
  dimLabels = c("grid", "plant")
)
```

---

# Results — grid level

## Observed interaction matrices

Each panel shows the sorted incidence/abundance matrix after SBM re-ordering,
highlighting the block structure.

```{r plot-observed, fig.width=14, fig.height=5}
gg_bact  <- plot_bipartite_sbm(sbm_Bact_gaussian)
gg_fungi <- plot_bipartite_sbm(sbm_Fungi_gaussian)
gg_virus <- plot_bipartite_sbm(sbm_Virus)
gg_plant <- plot_bipartite_sbm(sbm_Plant)

wrap_plots(list(gg_bact, gg_fungi, gg_virus, gg_plant), nrow = 1)
```

## Expected interaction matrices

The expected matrix (model reconstruction) smooths noise and makes the block
structure explicit.

```{r plot-expected, fig.width=28, fig.height=10}
gg_bact_exp  <- plot_bipartite_sbm(sbm_Bact_gaussian, mat = sbm_Bact_gaussian$expectation)
gg_fungi_exp <- plot_bipartite_sbm(sbm_Fungi_gaussian, mat = sbm_Fungi_gaussian$expectation)
gg_virus_exp <- plot_bipartite_sbm(sbm_Virus,          mat = sbm_Virus$expectation)
gg_plant_exp <- plot_bipartite_sbm(sbm_Plant,          mat = sbm_Plant$expectation)

wrap_plots(list(gg_bact_exp, gg_fungi_exp, gg_virus_exp, gg_plant_exp), nrow = 1)
```

## Average block signal per kingdom

```{r plot-avg-blocks, fig.width=10, fig.height=4}
sbm_Bact_avg  <- plot_bipartite_sbm_avg(sbm_Bact_gaussian)
sbm_Fungi_avg <- plot_bipartite_sbm_avg(sbm_Fungi_gaussian)
sbm_Virus_avg <- plot_bipartite_sbm_avg(sbm_Virus)

sbm_Bact_avg$plot
sbm_Fungi_avg$plot
sbm_Virus_avg$plot
```

## Cross-kingdom block co-membership (alluvial)

Each grid is assigned to one block per kingdom. The alluvial diagram shows how
the same grids cluster across bacteria, fungi, viruses and plants — revealing
whether ecological units are conserved or kingdom-specific.

```{r alluvial-plot, fig.width=10, fig.height=6}
# Sanity check: all SBMs operate on the same grid ordering
stopifnot(
  all(rownames(sbm_Bact_gaussian$networkData) == rownames(sbm_Fungi_gaussian$networkData)),
  all(rownames(sbm_Fungi_gaussian$networkData) == rownames(sbm_Virus$networkData)),
  all(rownames(sbm_Virus$networkData)          == rownames(sbm_Plant$networkData))
)

# Build contingency table of block co-memberships
B_multi_regne <- as.data.frame(table(
  grid  = rownames(sbm_Bact_gaussian$networkData),
  Bact  = sbm_Bact_gaussian$memberships$grid,
  Fungi = sbm_Fungi_gaussian$memberships$grid,
  Virus = sbm_Virus$memberships$grid,
  Plant = sbm_Plant$memberships$grid
)) %>%
  filter(Freq != 0)

alluvial(
  B_multi_regne[, c("Bact", "Fungi", "Virus", "Plant")],
  freq     = B_multi_regne$Freq,
  alpha    = 0.7,
  cex      = 1.5,
  cex.axis = 1.5,
  xw       = 0.2
)
```


## PLN — bloc-level JSDM (bacteria × fungi)

We use the SBM averaged expectations (`sbm_*_avg$matrix`, on the
`log1p(rarefied_count)` scale) as the per-bloc signal per grid. PLN is fit on
counts so we back-transform with `expm1()` and round to integers; this restores
expected per-bloc rarefied counts, which PLN's Poisson-lognormal handles cleanly
with `offset = "TSS"`.

Goal: partition joint bact × fungi variation into (a) environmental filtering
(β), (b) residual co-organization (Ω) after controlling for env, (c) noise.
PLN gives β + Ω directly; PLNnetwork gives sparse partial correlations; a
200-permutation null on rows of X tests inter-kingdom Ω entries.

### Build response Y

```{r pln-build-Y}
Y_bact  <- round(expm1(sbm_Bact_avg$matrix),  0)%>%
            as.data.frame()
Y_fungi <- round(expm1(sbm_Fungi_avg$matrix), 0)%>%
            as.data.frame()
Y_virus <- round(sbm_Virus_avg$matrix, 0)%>%
            as.data.frame()

Y_bact %>%
  rownames_to_column("grid") %>%
  left_join(Y_fungi %>%
            rownames_to_column("grid"),by = "grid") %>%
  left_join(Y_virus %>%
            rownames_to_column("grid"),by = "grid") %>%
  column_to_rownames("grid") %>%
  as.matrix() -> Y

storage.mode(Y) <- "integer"

kingdom <- c(rep("bact",  ncol(Y_bact)),
             rep("fungi", ncol(Y_fungi)),
             rep("virus", ncol(Y_virus)))
names(kingdom) <- colnames(Y)

cat("Y:", nrow(Y), "grids x", ncol(Y), "blocs",
    "(bact:", ncol(Y_bact), ", fungi:", ncol(Y_fungi), "virus:", ncol(Y_virus), ")\n")
cat("Min and max of Y:", range(Y), "\n")
```

### Build design matrix X

```{r pln-build-X}
meta <- read.table("data/bacteria/Metadata_grid_CAM.txt",
                   header = TRUE, stringsAsFactors = FALSE,
                   na.strings = c("NA", "", "."))

meta <- meta %>%
  filter(!str_detect(Grid_code, "21_CAM_1[45]")) 
rownames(Y)[!(rownames(Y) %in% meta$Grid_code)]

Y = Y[!(rownames(Y) == "22_CAM_15"),]
# Align meta rows to Y rows
stopifnot(all())
meta <- meta[match(rownames(Y), meta$Grid_code), , drop = FALSE]
stopifnot(identical(meta$Grid_code, rownames(Y)))

colnames(meta)
X <- meta %>%
  select(clust_landscp, clust_land_use, sbm_soil, clust_smb_grid_plant, 
         plant_richness, Biomass,plant_equit)
rownames(X) <- meta$Grid_code

cat("X:", nrow(X), "grids x", ncol(X), "covariates\n")
cat("n / p ratio:", round(nrow(X) / ncol(X), 1), "\n")
```

## Step 1 — Fit all models

We compare the 2³ design:

- **Model**: `full` (all covariates in `X`) vs `null` (intercept only)
- **Offset**: `off` (TSS library-size offset) vs `noff` (no offset)
- **Sparsity**: `pln` (dense PLN) vs `net` (PLNnetwork, BIC-selected best)

Total: 8 fits. All stored in named lists for easy access:
`fits_pln$<full|null>_<off|noff>`, `nets$<full|null>_<off|noff>`,
`bests$<full|null>_<off|noff>` (BIC-best PLNnetwork). `all_fits` is the
combined list keyed by `pln_*` / `net_*`.

To change the env covariates: edit the `pln-build-X` chunk above. The fitting
uses `~ .` so anything in `X` is picked up.

```{r pln-fit-all}

formulas <- list(
  full_off  = Abundance ~ . + offset(log(Offset)),
  full_noff = Abundance ~ .,
  null_off  = Abundance ~ 1 + offset(log(Offset)),
  null_noff = Abundance ~ 1
)

d        <- prepare_data(counts = Y, covariates = X, offset = "TSS")
fits_pln <- purrr::map(formulas, function(f) do.call(PLN,        list(formula = f, data = d)))
nets     <- purrr::map(formulas, function(f) do.call(PLNnetwork, list(formula = f, data = d)))


# BIC-best PLNnetwork models
bests <- purrr::map(nets, ~ getBestModel(.x, "BIC"))

# Combined list (8 entries) for downstream extraction
all_fits <- c(
  setNames(fits_pln, paste0("pln_", names(fits_pln))),
  setNames(bests,    paste0("net_", names(bests)))
)
names(all_fits)
```

## Step 2 — Per-model summary table

```{r pln-summary-table}
diag_df <- tibble::tibble(
  model_id = names(all_fits),
  loglik   = purrr::map_dbl(all_fits, "loglik"),
  BIC      = purrr::map_dbl(all_fits, "BIC")
)
knitr::kable(diag_df, digits = 1,
             caption = "Per-model log-likelihood and BIC (8 fits)")
```

## Step 3 — β extraction

Null models have no β (intercept only) and are skipped.

```{r pln-beta-extract}
extract_beta <- function(fit, label) {
  B <- coef(fit)
  if (is.null(B) || nrow(B) <= 1) return(tibble::tibble())   # null model
  B %>%
    as.data.frame() %>%
    rownames_to_column("covariate") %>%
   
    pivot_longer(-covariate, names_to = "bloc", values_to = "beta") %>%
    mutate(model_id = label)
}

beta_long <- purrr::imap_dfr(all_fits, extract_beta) %>%
  mutate(kingdom = kingdom[bloc],
         bloc    = factor(bloc, levels = colnames(Y)))
```

### β heatmap — env effects per bloc

```{r pln-beta-heatmap, fig.width=14, fig.height=8}
ggplot(beta_long, aes(x = bloc, y = covariate, fill = beta)) +
  geom_tile(colour = "white") +
  facet_grid(model_id ~ kingdom, scales = "free_x", space = "free_x") +
  scale_fill_gradient2(low = "#2c7bb6", mid = "white", high = "#d7191c",
                       name = expression(beta)) +
  labs(x = NULL, y = NULL,
       title = "β across full models (pln vs net × offset vs no-offset)") +
  theme_bw(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### β forest plot — distribution of effects per covariate

```{r pln-beta-forest, fig.width=12, fig.height=6}
ggplot(beta_long, aes(x = beta, y = covariate, colour = kingdom)) +
  geom_vline(xintercept = 0, linetype = 2, colour = "grey50") +
  geom_jitter(height = 0.15, alpha = 0.8, size = 1.5) +
  scale_colour_manual(values = c(bact = "#1b9e77", fungi = "#d95f02")) +
  facet_wrap(~ model_id, ncol = 2) +
  labs(x = expression(beta), y = NULL) +
  theme_bw(base_size = 10)
```

## Step 4 — Variance partitioning

Per (sparsity × offset): residual var = `diag(sigma(fit_full))`, total var =
`diag(sigma(fit_null))`, env share = `(total − residual) / total`. Four model
pairs.

```{r pln-vp-extract}
vp_pair <- function(fit_full, fit_null, label) {
  v_total <- diag(sigma(fit_null))
  v_resid <- diag(sigma(fit_full))
  v_env   <- pmax(v_total - v_resid, 0)
  tibble::tibble(
    bloc       = colnames(Y),
    kingdom    = kingdom,
    v_total    = v_total,
    v_env      = v_env,
    v_residual = v_resid,
    prop_env   = v_env / pmax(v_total, .Machine$double.eps),
    model_id   = label
  )
}

vp_long <- bind_rows(
  vp_pair(fits_pln$full_off,  fits_pln$null_off,  "pln_off"),
  vp_pair(fits_pln$full_noff, fits_pln$null_noff, "pln_noff"),
  vp_pair(bests$full_off,     bests$null_off,     "net_off"),
  vp_pair(bests$full_noff,    bests$null_noff,    "net_noff")
)

vp_summary <- vp_long %>%
  group_by(model_id, kingdom) %>%
  summarise(prop_env_mean = mean(prop_env),
            prop_env_sd   = sd(prop_env),
            n_blocs       = dplyr::n(),
            .groups = "drop")

knitr::kable(vp_summary, digits = 3,
             caption = "Variance partitioning summary (full vs null pairs)")
```

### Variance partitioning — stacked bar per bloc

```{r pln-vp-bar, fig.width=14, fig.height=8}
vp_long %>%
  pivot_longer(c(v_env, v_residual),
               names_to = "source", values_to = "var") %>%
  mutate(source = recode(source,
                         v_env      = "Environment",
                         v_residual = "Residual")) %>%
  ggplot(aes(x = bloc, y = var, fill = source)) +
  geom_col(position = "fill") +
  facet_grid(model_id ~ kingdom, scales = "free_x", space = "free_x") +
  scale_fill_manual(values = c("Environment" = "#fdae61",
                               "Residual"    = "#80b1d3")) +
  scale_y_continuous(labels = scales::percent) +
  labs(y = "Variance share", x = NULL, fill = NULL,
       title = "Variance partitioning per bloc, across models") +
  theme_bw(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Variance partitioning — per-kingdom boxplot

```{r pln-vp-boxplot, fig.width=12, fig.height=5}
ggplot(vp_long, aes(x = kingdom, y = prop_env, fill = kingdom)) +
  geom_boxplot(width = 0.4, outlier.shape = NA) +
  geom_jitter(width = 0.1, height = 0, alpha = 0.7) +
  facet_wrap(~ model_id, nrow = 1) +
  scale_fill_manual(values = c(bact = "#1b9e77", fungi = "#d95f02")) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(y = "Proportion of variance explained by env", x = NULL,
       title = "Environmental filtering, per kingdom, across models") +
  theme_bw(base_size = 10) +
  theme(legend.position = "none")
```

## Step 5 — Ω / partial-correlation heatmaps

Ω is the precision matrix (Σ⁻¹); partial correlations are `-Ω_{ij}/sqrt(Ω_{ii}·Ω_{jj})`.
For dense PLN fits Ω is obtained by inverting Σ; for PLNnetwork (BIC) best models
the sparse Ω is taken directly from the fitted model parameters.

```{r pln-omega-extract}
get_partial_cor <- function(fit) {
  Omega <- if (!is.null(fit$model_par$Omega)) fit$model_par$Omega else solve(sigma(fit))
  D     <- 1 / sqrt(diag(Omega))
  P     <- -sweep(sweep(Omega, 1, D, "*"), 2, D, "*")
  diag(P) <- 1
  P
}

omega_long <- purrr::imap_dfr(all_fits, function(fit, id) {
  M <- get_partial_cor(fit)
  dimnames(M) <- list(colnames(Y), colnames(Y))
  as.data.frame(M) %>%
    rownames_to_column("row") %>%
    pivot_longer(-row, names_to = "col", values_to = "value") %>%
    mutate(model_id = id)
})
```

### Ω heatmap grid (8 panels)

```{r pln-omega-heatmap, fig.width=16, fig.height=9, out.width="100%"}
n_blocs   <- length(bloc_order)
mid_bact  <- split_at / 2
mid_fungi <- split_at + (n_blocs - split_at) / 2

id_labels <- c(
  pln_full_off  = "PLN · Full + offset",
  pln_full_noff = "PLN · Full",
  pln_null_off  = "PLN · Intercept + offset",
  pln_null_noff = "PLN · Intercept only",
  net_full_off  = "PLNnet · Full + offset",
  net_full_noff = "PLNnet · Full",
  net_null_off  = "PLNnet · Intercept + offset",
  net_null_noff = "PLNnet · Intercept only"
)

omega_long %>%
  mutate(
    row      = factor(row, levels = bloc_order),
    col      = factor(col, levels = bloc_order),
    model_id = factor(model_id, levels = names(id_labels))
  ) %>%
  ggplot(aes(x = col, y = row, fill = value)) +
  geom_tile(colour = NA) +
  geom_hline(yintercept = split_at, colour = "grey10", linewidth = 1.2) +
  geom_vline(xintercept = split_at, colour = "grey10", linewidth = 1.2) +
  facet_wrap(~ model_id, nrow = 2, labeller = labeller(model_id = id_labels)) +
  scale_fill_gradient2(
    low    = "#2166ac",
    mid    = "white",
    high   = "#b2182b",
    limits = c(-1, 1),
    oob    = scales::squish,
    name   = "Partial\ncorrelation",
    breaks = c(-1, -0.5, 0, 0.5, 1)
  ) +
  scale_x_discrete(
    breaks = c(bloc_order[round(mid_bact)], bloc_order[round(mid_fungi)]),
    labels = c("Bacteria", "Fungi")
  ) +
  scale_y_discrete(
    breaks = c(bloc_order[round(mid_bact)], bloc_order[round(mid_fungi)]),
    labels = c("Bacteria", "Fungi")
  ) +
  coord_fixed(expand = FALSE) +
  labs(x = NULL, y = NULL) +
  theme_bw(base_size = 11) +
  theme(
    axis.text.x       = element_text(size = 9, colour = "grey30"),
    axis.text.y       = element_text(size = 9, colour = "grey30", angle = 90, hjust = 0.5),
    axis.ticks        = element_blank(),
    strip.text        = element_text(size = 9, face = "bold"),
    strip.background  = element_rect(fill = "#eef2f7", colour = "grey60", linewidth = 0.5),
    legend.position   = "right",
    legend.key.height = unit(3, "cm"),
    legend.key.width  = unit(0.5, "cm"),
    legend.title      = element_text(size = 9, face = "bold"),
    panel.border      = element_rect(colour = "grey60", linewidth = 0.5),
    panel.spacing     = unit(0.6, "lines"),
    plot.margin       = margin(5, 5, 5, 5)
  )
```

## Step 6 — PLNnetwork regularisation paths

```{r pln-network-path, fig.width=14, fig.height=4}
plots_path <- purrr::imap(nets, function(net, id) {
  plot(net, criteria = c("BIC", "ICL", "EBIC")) +
    labs(title = id)
})
patchwork::wrap_plots(plots_path, ncol = 4)
```

## Step 7 — Network plots (best PLNnetwork by BIC)

```{r pln-network-plot, fig.width=14, fig.height=4}
par(mfrow = c(1, 4))
purrr::iwalk(bests, function(best, id) plot(best, main = id))
par(mfrow = c(1, 1))
```

## Step 8 — Goodness of fit

Observed vs fitted on log1p scale, faceted by model. Null and no-offset
models will visibly deviate from the 1:1 line — that is the diagnostic.

```{r pln-goodness, fig.width=14, fig.height=8}
gof_long <- purrr::imap_dfr(all_fits, function(fit, id) {
  tibble::tibble(
    fitted   = as.vector(fitted(fit)),
    obs      = as.vector(Y),
    kingdom  = rep(kingdom, each = nrow(Y)),
    model_id = id
  )
})

ggplot(gof_long, aes(x = log1p(obs), y = log1p(fitted), colour = kingdom)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, colour = "grey50") +
  geom_point(alpha = 0.4) +
  facet_wrap(~ model_id, ncol = 4) +
  scale_colour_manual(values = c(bact = "#1b9e77", fungi = "#d95f02")) +
  labs(x = "log1p(observed)", y = "log1p(fitted)",
       title = "Goodness of fit across models") +
  theme_bw(base_size = 10)
```

<!-- ### Permutation null for inter-kingdom Ω entries -->

<!-- Permuting **rows of X** breaks any env-driven covariance between kingdoms while -->
<!-- preserving marginals and intra-kingdom signal. With n ≈ 42 and 200 refits this -->
<!-- takes a few minutes — cache the result. -->

<!-- ```{r pln-permutation} -->
<!-- inter_idx <- which(outer(kingdom, kingdom, "!="), arr.ind = TRUE) -->
<!-- inter_idx <- inter_idx[inter_idx[, 1] < inter_idx[, 2], , drop = FALSE] -->


<!-- set.seed(20260507) -->
<!-- n_perm    <- 1000 -->
<!-- perm_dist <- matrix(NA_real_, nrow = nrow(inter_idx), ncol = n_perm) -->

<!-- for (k in seq_len(n_perm)) { -->
<!--   perm_dist[, k] <- tryCatch({ -->
<!--     X_perm <- X[sample(nrow(X)), , drop = FALSE] -->
<!--     rownames(X_perm) <- rownames(X) -->
<!--     d_p    <- prepare_data(counts = Y, covariates = X_perm, offset = "TSS") -->
<!--     fit_p  <- PLN(Abundance ~ clust_landscp + clust_land_use + clust_land_use + -->
<!--                   clust_smb_grid_plant + plant_H + Biomass + virus_richness, data = d_p, -->
<!--                   control = PLN_param(trace = 0)) -->
<!--     sigma(fit_p)[inter_idx] -->
<!--   }, error = function(e) rep(NA_real_, nrow(inter_idx))) -->
<!-- } -->

<!-- obs_inter <- Omega[inter_idx] -->
<!-- p_inter   <- rowMeans(abs(perm_dist) >= abs(obs_inter), na.rm = TRUE) -->

<!-- str(perm_dist) -->
<!-- n_eff <- sum(!is.na(perm_dist[1, ])) -->
<!-- cat("Effective N permutations:", n_eff, "/", ncol(perm_dist), "\n") -->
<!-- cat("Inter-kingdom |Omega| entries with p < 0.05:", -->
<!--     sum(p_inter < 0.05, na.rm = TRUE), "/", length(p_inter), "\n") -->
<!-- ``` -->

<!-- ### Permutation null — top inter-kingdom Ω entries -->

<!-- ```{r pln-perm-histograms, fig.width=10, fig.height=7} -->
<!-- top_k_idx <- order(abs(Omega[inter_idx]), decreasing = TRUE)[ -->
<!--   seq_len(min(9, nrow(inter_idx))) -->
<!-- ] -->

<!-- pair_labels <- apply(inter_idx[top_k_idx, , drop = FALSE], 1, function(rc) { -->
<!--   sprintf("%s x %s", colnames(Y)[rc[1]], colnames(Y)[rc[2]]) -->
<!-- }) -->

<!-- perm_df <- purrr::map_dfr(seq_along(top_k_idx), function(j) { -->
<!--   k <- top_k_idx[j] -->
<!--   tibble( -->
<!--     label = pair_labels[j], -->
<!--     pval  = p_inter[k], -->
<!--     obs   = Omega[inter_idx][k], -->
<!--     perm  = perm_dist[k, ] -->
<!--   ) -->
<!-- }) %>% -->
<!--   mutate(facet_label = sprintf("%s\n(p = %.2f)", label, pval)) -->

<!-- ggplot(perm_df, aes(x = perm)) + -->
<!--   geom_histogram(bins = 30, fill = "grey80", colour = "grey40") + -->
<!--   geom_vline(aes(xintercept = obs), colour = "#d7191c", linewidth = 1) + -->
<!--   facet_wrap(~ facet_label, scales = "free") + -->
<!--   labs(x = "Omega (permuted)", y = "Count", -->
<!--        title = "Permutation null — top |Omega| inter-kingdom entries", -->
<!--        subtitle = "Red line = observed value") + -->
<!--   theme_bw(base_size = 10) -->
<!-- ``` -->

<!-- ### Ω heatmap with kingdom blocks and significance -->

<!-- Asterisks mark inter-kingdom entries with permutation p-value < 0.05. -->

<!-- ```{r pln-omega-heatmap, fig.width=10, fig.height=10} -->
<!-- sig_mat <- matrix("", nrow = nrow(Omega), ncol = ncol(Omega)) -->
<!-- for (k in seq_len(nrow(inter_idx))) { -->
<!--   i <- inter_idx[k, 1]; j <- inter_idx[k, 2] -->
<!--   if (!is.na(p_inter[k]) && p_inter[k] < 0.05) { -->
<!--     sig_mat[i, j] <- "*" -->
<!--     sig_mat[j, i] <- "*" -->
<!--   } -->
<!-- } -->
<!-- #BiocManager::install("ComplexHeatmap") -->

<!-- ComplexHeatmap::Heatmap( -->
<!--   Omega, -->
<!--   name            = "Omega", -->
<!--   cluster_rows    = FALSE, -->
<!--   cluster_columns = FALSE, -->
<!--   row_split       = factor(kingdom), -->
<!--   column_split    = factor(kingdom), -->
<!--   cell_fun = function(j, i, x, y, width, height, fill) { -->
<!--     grid::grid.text(sig_mat[i, j], x, y, gp = grid::gpar(fontsize = 9)) -->
<!--   } -->
<!-- ) -->

<!-- ComplexHeatmap::Heatmap( -->
<!--   Omega_sparse, -->
<!--   name            = "Omega", -->
<!--   cluster_rows    = FALSE, -->
<!--   cluster_columns = FALSE, -->
<!--   row_split       = factor(kingdom), -->
<!--   column_split    = factor(kingdom), -->
<!-- ) -->
<!-- ``` -->

---


# SBM at quadrat level (exploratory)

This section is work in progress. The goal is to run per-kingdom bipartite SBMs
at the finer quadrat scale (`YY_CAM_GG_QQ`) instead of the grid scale.

## Bacteria

### Data preparation

```{r quadra-bact-load}
# Retain only non-excluded quadrats, then truncate read depth (÷ 10)
ps_bacteria_quadra <- ps_bacteria %>%
  prune_samples(!str_detect(sample_names(.), "21-CAM-1[45]"), .)

cat("Quadrats available:", nsamples(ps_bacteria_quadra), "\n")

ps_bacteria_quadra %>%
  otu_table() %>%
  as.data.frame() %>%
  rename_with(~str_replace_all(., "-", "_")) %>%
  rename_with(~substring(., 1, 13)) -> bacteria_quadra_matrix

bacteria_quadra_truncated <- bacteria_quadra_matrix %>%
  rownames_to_column("OTU") %>%
  pivot_longer(-OTU, names_to = "quadra_code", values_to = "reads") %>%
  mutate(reads = floor(reads / 10)) %>%
  filter(reads > 0) %>%
  pivot_wider(names_from = "quadra_code", values_from = "reads", values_fill = 0) %>%
  column_to_rownames("OTU")

cat("After truncation:", nrow(bacteria_quadra_truncated), "OTUs x",
    ncol(bacteria_quadra_truncated), "quadrats\n")
```

### Read depth distribution

```{r quadra-bact-depth, fig.width=6, fig.height=4}
tibble(
  quadra  = colnames(bacteria_quadra_truncated),
  log10_reads = log10(colSums(bacteria_quadra_truncated))
) %>%
  arrange(log10_reads) %>%
  mutate(rank = row_number()) %>%
  ggplot(aes(rank, log10_reads)) +
  geom_point() +
  geom_hline(yintercept = log10(4), linetype = "dashed", colour = "red") +
  labs(x = "Quadrat rank", y = "log10(total reads)",
       title = "Read depth after truncation — quadrat level") +
  theme_bw()
```

### Low-coverage quadrat removal

Quadrats with sample coverage < 0.8 are removed before rarefaction to avoid
extrapolation artefacts.

```{r quadra-bact-coverage}
# Remove quadrats with fewer than 4 total reads after truncation
bacteria_quadra_truncated <- bacteria_quadra_truncated %>%
  rownames_to_column("OTU") %>%
  pivot_longer(-OTU, names_to = "quadra_code", values_to = "reads") %>%
  group_by(quadra_code) %>%
  filter(sum(reads) > 4) %>%
  ungroup() %>%
  pivot_wider(names_from = "quadra_code", values_from = "reads", values_fill = 0) %>%
  column_to_rownames("OTU")

iNext_bact_quadra <- bacteria_quadra_truncated %>%
  iNEXT3D(q = 0, datatype = "abundance", nboot = 0)

# Identify quadrats below coverage threshold
low_coverage_quadrats <- iNext_bact_quadra$TDiNextEst$coverage_based %>%
  filter(Order.q == 0, Method == "Observed", SC < 0.8) %>%
  pull(Assemblage)

cat("Quadrats removed (coverage < 0.8):", length(low_coverage_quadrats), "\n")

bacteria_quadra_filtered <- bacteria_quadra_truncated %>%
  rownames_to_column("OTU") %>%
  pivot_longer(-OTU, names_to = "quadra_code", values_to = "reads") %>%
  filter(!(quadra_code %in% low_coverage_quadrats), reads > 0) %>%
  pivot_wider(names_from = "quadra_code", values_from = "reads", values_fill = 0) %>%
  column_to_rownames("OTU")

cat("Remaining quadrats:", ncol(bacteria_quadra_filtered), "\n")
```

### Coverage-based rarefaction

```{r quadra-bact-rarefy-prep}
min_coverage_bact_quadra <- iNext_bact_quadra$TDiNextEst$coverage_based %>%
  filter(Order.q == 0, Method == "Observed", SC > 0.8) %>%
  slice_min(SC, n = 1) %>%
  pull(SC)

required_sizes_bact_quadra <- bacteria_quadra_filtered %>%
  estimate3D(level = min_coverage_bact_quadra, q = 0, nboot = 0) %>%
  select(Assemblage, m)

cat("Minimum coverage used for rarefaction:", round(min_coverage_bact_quadra, 3), "\n")
```

```{r quadra-bact-rarefy-sim, eval=FALSE}
n_sim  <- 100
set.seed(2365)

map(1:n_sim, ~coverage_rarefy_matrix(bacteria_quadra_filtered,
                                     required_sizes_bact_quadra)) -> quadra_bact_sim_list

quadra_bact_sim_list %>%
  map(~.x %>%
        as.data.frame() %>%
        rownames_to_column("OTU") %>%
        pivot_longer(-OTU, names_to = "sample", values_to = "count")) %>%
  reduce(full_join, by = c("OTU", "sample")) %>%
  mutate(total = round(rowSums(across(starts_with("count")), na.rm = TRUE) / n_sim)) %>%
  filter(total > 0) %>%
  select(OTU, sample, total) %>%
  pivot_wider(names_from = "sample", values_from = "total", values_fill = 0) %>%
  column_to_rownames("OTU") %>%
  as.matrix() -> quadra_bact_rarefied

cat("Retained OTUs:", nrow(quadra_bact_rarefied),
    "| Quadrats:", ncol(quadra_bact_rarefied), "\n")
```

### SBM fitting

```{r sbm-bact-quadra, eval=FALSE}
sbm_Bact_quadra_gaussian <- estimateBipartiteSBM(
  log1p(t(quadra_bact_rarefied)),
  model     = "gaussian",
  dimLabels = c("quadra", "bact")
)
save(sbm_Bact_quadra_gaussian,
     file = "outputs/heavy_computation_save/sbm_Bact_quadra_gaussian.Rdata")
```

```{r load-sbm-bact-quadra}
load("outputs/heavy_computation_save/sbm_Bact_quadra_gaussian.Rdata")
plot_bipartite_sbm(sbm_Bact_quadra_gaussian)
```

## Fungi (work in progress)

Fungi at the quadrat level are truncated with a divisor of 5 (shallower
sequencing than bacteria) before coverage estimation.

```{r quadra-fungi-prep}
ps_fungi_quadra <- ps_fungi %>%
  prune_samples(!str_detect(sample_names(.), "21-CAM-1[45]"), .)

ps_fungi_quadra %>%
  otu_table() %>%
  as.data.frame() %>%
  rename_with(~str_replace_all(., "-", "_")) -> fungi_quadra_matrix

fungi_quadra_truncated <- fungi_quadra_matrix %>%
  rownames_to_column("OTU") %>%
  pivot_longer(-OTU, names_to = "quadra_code", values_to = "reads") %>%
  mutate(reads = floor(reads / 5)) %>%
  filter(reads > 0) %>%
  pivot_wider(names_from = "quadra_code", values_from = "reads", values_fill = 0) %>%
  column_to_rownames("OTU")

cat("Fungi quadrat matrix after truncation:", nrow(fungi_quadra_truncated), "OTUs x",
    ncol(fungi_quadra_truncated), "quadrats\n")
```

> **Note:** Coverage estimation and SBM fitting for fungi at the quadrat level
> mirror the bacteria workflow above and are pending implementation.
