---
title: "Bipartite SBM per Kingdom"
author: "Mathis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  cache   = FALSE
)
```

# Packages and helpers

```{r packages}
library(tidyverse)
library(phyloseq)
library(sbm)
library(iNEXT.3D)
library(patchwork)
library(alluvial)

source("scripts/plot_multipartite_sbm.R")
source("scripts/functions_modified.R")
```

---

# Data loading

All four kingdoms are aggregated to the **grid level** (sampling unit: `YY_CAM_GG`).
Grids 14 and 15 from 2021 are excluded throughout.

## Bacteria

```{r load-bacteria}
load("outputs/clean_data_to_analise/ps_bacteria.Rdata")

ps_bacteria_grid <- merge_samples(ps_bacteria, "grid") %>%
  prune_samples(!str_detect(sample_names(.), "21-CAM-1[45]"), .)

# OTU table: rows = OTUs, columns = grids
ps_bacteria_grid %>%
  otu_table() %>%
  t() %>%
  as.data.frame() %>%
  rename_with(~str_replace_all(., "-", "_")) -> bacteria_grid_matrix

# The column order of bacteria defines the reference sort for all other kingdoms
reference_sample_sort <- colnames(bacteria_grid_matrix)

cat("Bacteria grid matrix:", nrow(bacteria_grid_matrix), "OTUs x", ncol(bacteria_grid_matrix), "grids\n")
```

## Fungi

```{r load-fungi}
load("outputs/clean_data_to_analise/ps_fungi.Rdata")

ps_fungi_grid <- merge_samples(ps_fungi, "grid_code") %>%
  prune_samples(!str_detect(sample_names(.), "21_CAM_1[45]"), .)

# Align columns to bacteria sort order
ps_fungi_grid %>%
  otu_table() %>%
  t() %>%
  as.data.frame() -> fungi_grid_matrix

fungi_grid_matrix <- fungi_grid_matrix[, reference_sample_sort]

cat("Fungi grid matrix:", nrow(fungi_grid_matrix), "OTUs x", ncol(fungi_grid_matrix), "grids\n")
```

## Viruses

Virus data are at quadrat level; they are summed to grid level here.

```{r load-virus}
read.table("outputs/clean_data_to_analise/OTU_virus_CAM.txt", header = TRUE) %>%
  mutate(grid_code = str_sub(Host_code, 1, 9)) %>%
  filter(!str_detect(grid_code, "21_CAM_1[45]")) %>%
  select(-Host_code) %>%
  group_by(grid_code) %>%
  summarise(across(where(is.numeric), sum)) %>%
  column_to_rownames("grid_code") %>%
  as.matrix() -> virus_grid_matrix

# Align rows to reference sort
virus_grid_matrix <- virus_grid_matrix[reference_sample_sort, ]

cat("Virus grid matrix:", nrow(virus_grid_matrix), "grids x", ncol(virus_grid_matrix), "OTUs\n")
```

## Plants

Plant abundances are already aggregated at grid level (quadrat cover × 2, summed).

```{r load-plants}
read.table("outputs/clean_data_to_analise/abund_plant_grid.txt", header = TRUE) %>%
  column_to_rownames("Grid_code") %>%
  as.matrix() -> plant_grid_matrix

plant_grid_matrix <- plant_grid_matrix[reference_sample_sort, ]

cat("Plant grid matrix:", nrow(plant_grid_matrix), "grids x", ncol(plant_grid_matrix), "OTUs\n")
```

## Metadata

```{r load-metadata}
read.table("outputs/clean_data_to_analise/Metadata_grid_CAM.txt", header = TRUE) -> metadata_grid
cat("Metadata:", nrow(metadata_grid), "grids,", ncol(metadata_grid), "variables\n")
```

---

# SBM at grid level

## Read count truncation

High sequencing depth inflates the matrix and slows SBM fitting.
Reads are floored at 1/10 of their original value; zero-count entries are then
dropped to keep the matrix sparse.

```{r truncate-grid}
truncate_matrix <- function(otu_mat, divisor = 10) {
  otu_mat %>%
    rownames_to_column("OTU") %>%
    pivot_longer(-OTU, names_to = "sample", values_to = "reads") %>%
    mutate(reads = floor(reads / divisor)) %>%
    filter(reads > 0) %>%
    pivot_wider(names_from = "sample", values_from = "reads", values_fill = 0) %>%
    column_to_rownames("OTU")
}

bacteria_grid_truncated <- truncate_matrix(bacteria_grid_matrix)
fungi_grid_truncated    <- truncate_matrix(fungi_grid_matrix)

cat("After truncation — Bacteria:", nrow(bacteria_grid_truncated), "OTUs\n")
cat("After truncation — Fungi   :", nrow(fungi_grid_truncated),    "OTUs\n")
```

## Coverage-based rarefaction

### Minimum observed coverage per kingdom

All samples within a kingdom are rarefied to the same coverage level — the
minimum observed sample coverage — so that richness comparisons are not biased
by sequencing depth.

```{r coverage-bacteria}
iNext_bact <- bacteria_grid_truncated %>%
  iNEXT3D(q = 0, datatype = "abundance", nboot = 0)

min_coverage_bact <- iNext_bact$TDiNextEst$coverage_based %>%
  filter(Order.q == 0, Method == "Observed") %>%
  slice_min(SC, n = 1) %>%
  pull(SC)

required_sizes_bact <- bacteria_grid_truncated %>%
  estimate3D(level = min_coverage_bact, q = 0, nboot = 0) %>%
  select(Assemblage, m)

cat("Minimum sample coverage (bacteria):", round(min_coverage_bact, 3), "\n")
```

```{r coverage-fungi}
iNext_fungi <- fungi_grid_truncated %>%
  iNEXT3D(q = 0, datatype = "abundance", nboot = 0)

min_coverage_fungi <- iNext_fungi$TDiNextEst$coverage_based %>%
  filter(Order.q == 0, Method == "Observed") %>%
  slice_min(SC, n = 1) %>%
  pull(SC)

required_sizes_fungi <- fungi_grid_truncated %>%
  estimate3D(level = min_coverage_fungi, q = 0, nboot = 0) %>%
  select(Assemblage, m)

cat("Minimum sample coverage (fungi):", round(min_coverage_fungi, 3), "\n")
```

Comparison of target rarefaction sizes between bacteria and fungi:

```{r coverage-comparison-plot, fig.width=5, fig.height=4}
required_sizes_fungi %>%
  left_join(required_sizes_bact, by = "Assemblage", suffix = c("_fungi", "_bact")) %>%
  ggplot(aes(m_fungi, m_bact)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Target size (fungi)", y = "Target size (bacteria)") +
  theme_bw()
```

### Rarefaction function

The function samples `target_size` reads without replacement from each sample,
repeating the process `n_sim` times. The per-OTU totals across simulations are
then accumulated into a single matrix (effectively averaging over stochasticity).

```{r rarefy-function}
coverage_rarefy_matrix <- function(otu_matrix, required_sizes) {
  sapply(colnames(otu_matrix), function(sample_name) {
    sample_vector <- otu_matrix[, sample_name]
    target_size <- required_sizes %>%
      filter(Assemblage == sample_name) %>%
      pull(m)

    # Pool all reads as individual OTU tags, then draw without replacement
    otu_pool    <- rep(rownames(otu_matrix), times = sample_vector)
    sampled     <- sample(otu_pool, size = target_size, replace = FALSE)
    rarefied    <- table(factor(sampled, levels = rownames(otu_matrix)))
    as.numeric(rarefied)
  }) %>%
    `rownames<-`(rownames(otu_matrix)) %>%
    as.data.frame() %>%
    filter(rowSums(across(where(is.numeric))) != 0) %>%
    as.matrix()
}
```

### Simulation (n = 100)

Each simulation draws a new rarefied matrix; the 100 matrices are stacked and
summed per (OTU, sample) cell. OTUs present in fewer than 4 reads on average
(i.e. total < 400 across simulations) are discarded as likely noise.

```{r rarefy-grid, eval=FALSE}
n_sim  <- 100
set.seed(2365)

accumulate_simulations <- function(sim_list, n_sim) {
  sim_list %>%
    map(~.x %>%
          as.data.frame() %>%
          rownames_to_column("OTU") %>%
          pivot_longer(-OTU, names_to = "sample", values_to = "count")) %>%
    reduce(full_join, by = c("OTU", "sample")) %>%
    mutate(total = rowSums(across(starts_with("count")), na.rm = TRUE)) %>%
    select(OTU, sample, total) %>%
    pivot_wider(names_from = "sample", values_from = "total") %>%
    column_to_rownames("OTU") %>%
    as.matrix()
}

# Bacteria
map(1:n_sim, ~coverage_rarefy_matrix(bacteria_grid_truncated,
                                     required_sizes_bact)) -> bact_sim_list
bact_grid_cumul <- accumulate_simulations(bact_sim_list, n_sim)
# Keep OTUs with mean abundance > 3 (total > 3 * n_sim)
bact_grid_rarefied <- bact_grid_cumul[rowSums(round(bact_grid_cumul / n_sim)) > 3, ]

# Fungi
map(1:n_sim, ~coverage_rarefy_matrix(fungi_grid_truncated,
                                     required_sizes_fungi)) -> fungi_sim_list
fungi_grid_cumul <- accumulate_simulations(fungi_sim_list, n_sim)
fungi_grid_rarefied <- fungi_grid_cumul[rowSums(round(fungi_grid_cumul / n_sim)) > 3, ]

cat("Retained OTUs — Bacteria:", nrow(bact_grid_rarefied),
    "| Fungi:", nrow(fungi_grid_rarefied), "\n")

# Sanity check: grid order must match across all kingdoms
stopifnot(
  all(colnames(bact_grid_rarefied) == colnames(fungi_grid_rarefied)),
  all(colnames(fungi_grid_rarefied) == rownames(virus_grid_matrix)),
  all(rownames(virus_grid_matrix)   == rownames(plant_grid_matrix))
)
```

## SBM fitting

Bipartite SBMs relate grids (row nodes) to OTUs (column nodes). The emission
model is chosen per kingdom:

- **Bacteria / Fungi**: Gaussian on log1p-transformed cumulated rarefied counts.
- **Virus**: Poisson on raw counts (low read depth, integers well-behaved).
- **Plants**: Gaussian on log1p-transformed abundance.

Heavy computation blocks below are set to `eval = FALSE`; cached objects are
loaded immediately after.

### Bacteria

```{r sbm-bacteria, eval=FALSE}
sbm_Bact_gaussian <- estimateBipartiteSBM(
  log1p(t(bact_grid_rarefied)),
  model     = "gaussian",
  dimLabels = c("grid", "bact")
)
save(sbm_Bact_gaussian, file = "outputs/heavy_computation_save/sbm_Bact_gaussian.Rdata")
```

```{r load-sbm-bacteria}
load("outputs/heavy_computation_save/sbm_Bact_gaussian.Rdata")
```

### Fungi

```{r sbm-fungi, eval=FALSE}
sbm_Fungi_gaussian <- estimateBipartiteSBM(
  log1p(t(fungi_grid_rarefied)),
  model     = "gaussian",
  dimLabels = c("grid", "fungi")
)
save(sbm_Fungi_gaussian, file = "outputs/heavy_computation_save/sbm_Fungi_gaussian.Rdata")
```

```{r load-sbm-fungi}
load("outputs/heavy_computation_save/sbm_Fungi_gaussian.Rdata")
```

### Viruses

```{r sbm-virus}
sbm_Virus <- estimateBipartiteSBM(
  virus_grid_matrix,
  model     = "poisson",
  dimLabels = c("grid", "virus")
)
```

### Plants

```{r sbm-plants}
sbm_Plant <- estimateBipartiteSBM(
  log1p(plant_grid_matrix),
  model     = "gaussian",
  dimLabels = c("grid", "plant")
)
```

---

# Results — grid level

## Observed interaction matrices

Each panel shows the sorted incidence/abundance matrix after SBM re-ordering,
highlighting the block structure.

```{r plot-observed, fig.width=14, fig.height=5}
gg_bact  <- plot_bipartite_sbm(sbm_Bact_gaussian)
gg_fungi <- plot_bipartite_sbm(sbm_Fungi_gaussian)
gg_virus <- plot_bipartite_sbm(sbm_Virus)
gg_plant <- plot_bipartite_sbm(sbm_Plant)

wrap_plots(list(gg_bact, gg_fungi, gg_virus, gg_plant), nrow = 1)
```

## Expected interaction matrices

The expected matrix (model reconstruction) smooths noise and makes the block
structure explicit.

```{r plot-expected, fig.width=14, fig.height=5}
gg_bact_exp  <- plot_bipartite_sbm(sbm_Bact_gaussian, mat = sbm_Bact_gaussian$expectation)
gg_fungi_exp <- plot_bipartite_sbm(sbm_Fungi_gaussian, mat = sbm_Fungi_gaussian$expectation)
gg_virus_exp <- plot_bipartite_sbm(sbm_Virus,          mat = sbm_Virus$expectation)
gg_plant_exp <- plot_bipartite_sbm(sbm_Plant,          mat = sbm_Plant$expectation)

wrap_plots(list(gg_bact_exp, gg_fungi_exp, gg_virus_exp, gg_plant_exp), nrow = 1)
```

## Average block signal per kingdom

```{r plot-avg-blocks, fig.width=10, fig.height=4}
sbm_Bact_avg  <- plot_bipartite_sbm_avg(sbm_Bact_gaussian)
sbm_Fungi_avg <- plot_bipartite_sbm_avg(sbm_Fungi_gaussian)

sbm_Bact_avg$plot
sbm_Fungi_avg$plot
```

## Cross-kingdom block co-membership (alluvial)

Each grid is assigned to one block per kingdom. The alluvial diagram shows how
the same grids cluster across bacteria, fungi, viruses and plants — revealing
whether ecological units are conserved or kingdom-specific.

```{r alluvial-plot, fig.width=10, fig.height=6}
# Sanity check: all SBMs operate on the same grid ordering
stopifnot(
  all(rownames(sbm_Bact_gaussian$networkData) == rownames(sbm_Fungi_gaussian$networkData)),
  all(rownames(sbm_Fungi_gaussian$networkData) == rownames(sbm_Virus$networkData)),
  all(rownames(sbm_Virus$networkData)          == rownames(sbm_Plant$networkData))
)

# Build contingency table of block co-memberships
B_multi_regne <- as.data.frame(table(
  grid  = rownames(sbm_Bact_gaussian$networkData),
  Bact  = sbm_Bact_gaussian$memberships$grid,
  Fungi = sbm_Fungi_gaussian$memberships$grid,
  Virus = sbm_Virus$memberships$grid,
  Plant = sbm_Plant$memberships$grid
)) %>%
  filter(Freq != 0)

alluvial(
  B_multi_regne[, c("Bact", "Fungi", "Virus", "Plant")],
  freq     = B_multi_regne$Freq,
  alpha    = 0.7,
  cex      = 1.5,
  cex.axis = 1.5,
  xw       = 0.2
)
```

---

# SBM at quadrat level (exploratory)

This section is work in progress. The goal is to run per-kingdom bipartite SBMs
at the finer quadrat scale (`YY_CAM_GG_QQ`) instead of the grid scale.

## Bacteria

### Data preparation

```{r quadra-bact-load}
# Retain only non-excluded quadrats, then truncate read depth (÷ 10)
ps_bacteria_quadra <- ps_bacteria %>%
  prune_samples(!str_detect(sample_names(.), "21-CAM-1[45]"), .)

cat("Quadrats available:", nsamples(ps_bacteria_quadra), "\n")

ps_bacteria_quadra %>%
  otu_table() %>%
  as.data.frame() %>%
  rename_with(~str_replace_all(., "-", "_")) %>%
  rename_with(~substring(., 1, 13)) -> bacteria_quadra_matrix

bacteria_quadra_truncated <- bacteria_quadra_matrix %>%
  rownames_to_column("OTU") %>%
  pivot_longer(-OTU, names_to = "quadra_code", values_to = "reads") %>%
  mutate(reads = floor(reads / 10)) %>%
  filter(reads > 0) %>%
  pivot_wider(names_from = "quadra_code", values_from = "reads", values_fill = 0) %>%
  column_to_rownames("OTU")

cat("After truncation:", nrow(bacteria_quadra_truncated), "OTUs x",
    ncol(bacteria_quadra_truncated), "quadrats\n")
```

### Read depth distribution

```{r quadra-bact-depth, fig.width=6, fig.height=4}
tibble(
  quadra  = colnames(bacteria_quadra_truncated),
  log10_reads = log10(colSums(bacteria_quadra_truncated))
) %>%
  arrange(log10_reads) %>%
  mutate(rank = row_number()) %>%
  ggplot(aes(rank, log10_reads)) +
  geom_point() +
  geom_hline(yintercept = log10(4), linetype = "dashed", colour = "red") +
  labs(x = "Quadrat rank", y = "log10(total reads)",
       title = "Read depth after truncation — quadrat level") +
  theme_bw()
```

### Low-coverage quadrat removal

Quadrats with sample coverage < 0.8 are removed before rarefaction to avoid
extrapolation artefacts.

```{r quadra-bact-coverage}
# Remove quadrats with fewer than 4 total reads after truncation
bacteria_quadra_truncated <- bacteria_quadra_truncated %>%
  rownames_to_column("OTU") %>%
  pivot_longer(-OTU, names_to = "quadra_code", values_to = "reads") %>%
  group_by(quadra_code) %>%
  filter(sum(reads) > 4) %>%
  ungroup() %>%
  pivot_wider(names_from = "quadra_code", values_from = "reads", values_fill = 0) %>%
  column_to_rownames("OTU")

iNext_bact_quadra <- bacteria_quadra_truncated %>%
  iNEXT3D(q = 0, datatype = "abundance", nboot = 0)

# Identify quadrats below coverage threshold
low_coverage_quadrats <- iNext_bact_quadra$TDiNextEst$coverage_based %>%
  filter(Order.q == 0, Method == "Observed", SC < 0.8) %>%
  pull(Assemblage)

cat("Quadrats removed (coverage < 0.8):", length(low_coverage_quadrats), "\n")

bacteria_quadra_filtered <- bacteria_quadra_truncated %>%
  rownames_to_column("OTU") %>%
  pivot_longer(-OTU, names_to = "quadra_code", values_to = "reads") %>%
  filter(!(quadra_code %in% low_coverage_quadrats), reads > 0) %>%
  pivot_wider(names_from = "quadra_code", values_from = "reads", values_fill = 0) %>%
  column_to_rownames("OTU")

cat("Remaining quadrats:", ncol(bacteria_quadra_filtered), "\n")
```

### Coverage-based rarefaction

```{r quadra-bact-rarefy-prep}
min_coverage_bact_quadra <- iNext_bact_quadra$TDiNextEst$coverage_based %>%
  filter(Order.q == 0, Method == "Observed", SC > 0.8) %>%
  slice_min(SC, n = 1) %>%
  pull(SC)

required_sizes_bact_quadra <- bacteria_quadra_filtered %>%
  estimate3D(level = min_coverage_bact_quadra, q = 0, nboot = 0) %>%
  select(Assemblage, m)

cat("Minimum coverage used for rarefaction:", round(min_coverage_bact_quadra, 3), "\n")
```

```{r quadra-bact-rarefy-sim, eval=FALSE}
n_sim  <- 100
set.seed(2365)

map(1:n_sim, ~coverage_rarefy_matrix(bacteria_quadra_filtered,
                                     required_sizes_bact_quadra)) -> quadra_bact_sim_list

quadra_bact_sim_list %>%
  map(~.x %>%
        as.data.frame() %>%
        rownames_to_column("OTU") %>%
        pivot_longer(-OTU, names_to = "sample", values_to = "count")) %>%
  reduce(full_join, by = c("OTU", "sample")) %>%
  mutate(total = round(rowSums(across(starts_with("count")), na.rm = TRUE) / n_sim)) %>%
  filter(total > 0) %>%
  select(OTU, sample, total) %>%
  pivot_wider(names_from = "sample", values_from = "total", values_fill = 0) %>%
  column_to_rownames("OTU") %>%
  as.matrix() -> quadra_bact_rarefied

cat("Retained OTUs:", nrow(quadra_bact_rarefied),
    "| Quadrats:", ncol(quadra_bact_rarefied), "\n")
```

### SBM fitting

```{r sbm-bact-quadra, eval=FALSE}
sbm_Bact_quadra_gaussian <- estimateBipartiteSBM(
  log1p(t(quadra_bact_rarefied)),
  model     = "gaussian",
  dimLabels = c("quadra", "bact")
)
save(sbm_Bact_quadra_gaussian,
     file = "outputs/heavy_computation_save/sbm_Bact_quadra_gaussian.Rdata")
```

```{r load-sbm-bact-quadra}
load("outputs/heavy_computation_save/sbm_Bact_quadra_gaussian.Rdata")
plot_bipartite_sbm(sbm_Bact_quadra_gaussian)
```

## Fungi (work in progress)

Fungi at the quadrat level are truncated with a divisor of 5 (shallower
sequencing than bacteria) before coverage estimation.

```{r quadra-fungi-prep}
ps_fungi_quadra <- ps_fungi %>%
  prune_samples(!str_detect(sample_names(.), "21-CAM-1[45]"), .)

ps_fungi_quadra %>%
  otu_table() %>%
  as.data.frame() %>%
  rename_with(~str_replace_all(., "-", "_")) -> fungi_quadra_matrix

fungi_quadra_truncated <- fungi_quadra_matrix %>%
  rownames_to_column("OTU") %>%
  pivot_longer(-OTU, names_to = "quadra_code", values_to = "reads") %>%
  mutate(reads = floor(reads / 5)) %>%
  filter(reads > 0) %>%
  pivot_wider(names_from = "quadra_code", values_from = "reads", values_fill = 0) %>%
  column_to_rownames("OTU")

cat("Fungi quadrat matrix after truncation:", nrow(fungi_quadra_truncated), "OTUs x",
    ncol(fungi_quadra_truncated), "quadrats\n")
```

> **Note:** Coverage estimation and SBM fitting for fungi at the quadrat level
> mirror the bacteria workflow above and are pending implementation.
