---
title: "Untitled"
output: html_document
date: "2026-03-18"
editor_options: 
  chunk_output_type: inline
---
# Load package and data  
```{r}
# if (!require("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# 
# BiocManager::install("phyloseq")

library(tidyverse)
library(phyloseq)
library(iNEXT.3D)
library(vegan)
library(alluvial)
library(viridis)
library(sbm)
library(furrr)
```

```{r}
load(file = "outputs/clean_data_to_analise/ps_bacteria.Rdata")
```

# OTU level coverage rarefation and SBM

```{r}
coverage_rarefy_matrix <- function(otu_matrix, 
                                   required_sizes,
                                   drop_lowcoverage = FALSE){
  
  # Calculer le coverage de chaque échantillon
  sample_coverages <- apply(otu_matrix, 2, function(sample_vector) {
    sample_vector <- sample_vector[sample_vector > 0]
    n <- sum(sample_vector)
    f1 <- sum(sample_vector == 1)
    f2 <- sum(sample_vector == 2)
    
    if (f1 > 0 && f2 > 0) {
      1 - (f1 / n) * (((n - 1) * f1) / ((n - 1) * f1 + 2 * f2))
    } else if (f1 > 0 && f2 == 0) {
      1 - (f1 / n) * (((n - 1) * f1) / ((n - 1) * f1 + 2))
    } else {
      1
    }
  })
  
  # Raréfier chaque échantillon
  rarefied_matrix <- sapply(colnames(otu_matrix), function(sample_name) {
    sample_vector <- otu_matrix[, sample_name]
    target_size <- required_sizes %>%
      filter(Assemblage == sample_name) %>%
      pull(m)
    
    # Créer un vecteur répété de chaque OTU selon son abondance
    otu_pool <- rep(rownames(otu_matrix), times = sample_vector)
    
    # Échantillonner sans remise
    sampled_otus <- sample(otu_pool, size = target_size, replace = FALSE)
    
    # Compter les abondances
    rarefied <- table(factor(sampled_otus, levels = rownames(otu_matrix)))
    
    return(as.numeric(rarefied))
  })
  
  rownames(rarefied_matrix) <- rownames(otu_matrix)
  rarefied_matrix %>%
    as.data.frame() %>%
    filter(rowSums(across(where(is.numeric)))!=0) %>%
    as.matrix() -> rarefied_matrix
  
  return(rarefied_matrix)
}
```

```{r}
calculate.coverage.f1<- function(f1,f2, sample_size){1-(f1/sample_size) * ((sample_size-1)*f1)/((sample_size-1)*f1 + 2*f2)}

ps_bacteria %>%
  otu_table() %>%
  as.data.frame() %>%
  rownames_to_column("OTU") %>%
  pivot_longer(-OTU, names_to = "quadrat_code", values_to = "reads") %>%
  mutate(grid_code = str_sub(quadrat_code,1,9),
         reads = floor(reads/10)) %>%
  filter(reads > 0) -> trucated_bacteria_quadat

trucated_bacteria_quadat %>%
  select(-quadrat_code) %>%
  group_by(grid_code,OTU) %>%
  summarise(reads = sum(reads))  -> trucated_bacteria_grid
```

```{r}
trucated_bacteria_quadat %>%
  group_by(quadrat_code) %>%
  mutate(sample_size =sum(reads)) %>%
  filter(reads <= 4, reads!=0) %>% 
  group_by(quadrat_code, reads, sample_size) %>% 
  summarize(count = n(), .groups = "drop") %>% 
  pivot_wider(names_from =reads , values_from = count, values_fill = 0) %>%
  rename(f1 ="1", f2 ="2", f3 ="3", f4 ="4") %>%
  mutate(f1_hat = ifelse(f3*f4>0, 4*f2^2/(3*f3)-f2*f3/(2*f4), 4*f2^2/(3*(f3+1))-f2*f3/(2*(f4+1))),
         coverage = calculate.coverage.f1(f1 ,f2 , sample_size),
         coverage_hat = calculate.coverage.f1(f1_hat, f2, sample_size)) -> quadrat_coverage_df

trucated_bacteria_grid %>%
  group_by(grid_code) %>%
  mutate(sample_size =sum(reads)) %>%
  filter(reads <= 4, reads!=0) %>% 
  group_by(grid_code, reads, sample_size) %>% 
  summarize(count = n(), .groups = "drop") %>% 
  pivot_wider(names_from =reads , values_from = count, values_fill = 0) %>%
  rename(f1 ="1", f2 ="2", f3 ="3", f4 ="4") %>%
  mutate(f1_hat = ifelse(f3*f4>0, 4*f2^2/(3*f3)-f2*f3/(2*f4), 4*f2^2/(3*(f3+1))-f2*f3/(2*(f4+1))),
         coverage = calculate.coverage.f1(f1 ,f2 , sample_size),
         coverage_hat = calculate.coverage.f1(f1_hat, f2, sample_size)) -> grid_coverage_df

quadrat_coverage_df
grid_coverage_df
```

## Find samples sizes (at reference coverage)

```{r}
trucated_bacteria_grid %>%
              pivot_wider(names_from = "grid_code", values_from = "reads", values_fill = 0) %>%
              column_to_rownames("OTU") -> OTU_matrix_trucated

str(as.matrix(OTU_matrix_trucated))
# Calculer les tailles requises
iNext_object_OTU <- OTU_matrix_trucated %>%
  iNEXT.3D::iNEXT3D( q = c(0), datatype = "abundance", nboot = 0)

iNext_object_OTU$TDiNextEst$coverage_based %>% 
  filter(Order.q == 0 & Method == "Observed") %>%
  filter(SC == min(SC, na.rm = TRUE)) %>%
  pull(SC) -> min_coverage_OTU
grid_coverage_df %>%
    filter(coverage_hat == min(coverage_hat, na.rm = TRUE)) %>%
    pull(coverage_hat) 

OTU_matrix_trucated %>%
  estimate3D(level = min_coverage_OTU, q = 0, nboot = 0) %>%
  select(Assemblage,m) -> required_sizes_OTU
```

## Rarefaction at reference coverage

```{r}
n_sim = 100
set.seed(2365)
1:n_sim %>%
  map(\(x) OTU_matrix_trucated %>%
              coverage_rarefy_matrix(required_sizes = required_sizes_OTU)) -> coverage_rarefy_OTU_list
```

```{r}
coverage_rarefy_OTU_list[[1]] %>% str
coverage_rarefy_OTU_list[[2]] %>% str
```

```{r}
if(file.exists("outputs/heavy_computation_save/coverage_subsample_sbm_bact.Rdata")){

  load(file = "outputs/heavy_computation_save/coverage_subsample_sbm_bact.Rdata")

  }else{
    
  plan(multisession, workers = 12)
  coverage_rarefy_OTU_list %>%
    furrr::future_map(\(x) x%>% 
                        as.matrix() %>%
                        log1p %>% 
                        estimateBipartiteSBM(netMat= .,
                                             model = "poisson",
                                             dimLabels = c(row = "OTU", col = "grid_code"),
                                             estimOptions = list(plot = FALSE,
                                                                 verbosity = 0)),
                      .options = furrr_options(scheduling = 2),
                      .progress = T
  ) -> coverage_subsample_OTU_sbm_list

save(coverage_subsample_OTU_sbm_list, file = "outputs/heavy_computation_save/coverage_subsample_sbm_bact.Rdata")
  }
```

```{r}
map(1:n_sim, \(x) data.frame(subsample = coverage_subsample_OTU_sbm_list[[x]]$memberships$grid_code,
                             grid_code = colnames(coverage_rarefy_OTU_list[[1]]),
                             sim = x)
    ) %>%
  bind_rows() %>%
  pivot_wider(names_from = sim, names_glue = "simulation{sim}", values_from = "subsample") %>%
  column_to_rownames("grid_code") -> cluster_coverage_subsample_df
```

```{r}
n_echantillons <- nrow(cluster_coverage_subsample_df)
coverage_subsample_cluster_mat <- matrix(0, nrow = n_echantillons, ncol = n_echantillons)

for (sim in 1:n_sim) {
  coverage_subsample_cluster_mat <- coverage_subsample_cluster_mat + outer(cluster_coverage_subsample_df[[sim]], cluster_coverage_subsample_df[[sim]], "==")
}

rownames(coverage_subsample_cluster_mat) <- rownames(cluster_coverage_subsample_df)
colnames(coverage_subsample_cluster_mat) <- rownames(cluster_coverage_subsample_df)
coverage_subsample_cluster_mat %>%
  as.data.frame() %>%
  rownames_to_column("grid_code") %>%
  pivot_longer(-grid_code, names_to = "name", values_to = "value") %>%
  # Create two temporary columns where the names are ALWAYS sorted
  # pmin/pmax ensures that (Sample_A, Sample_B) and (Sample_B, Sample_A) 
  # both become (Sample_A, Sample_B)
  mutate(
    p1 = pmin(grid_code, name),
    p2 = pmax(grid_code, name)
  ) %>%
  # Distinct now recognizes these rows as identical
  distinct(p1, p2, .keep_all = TRUE) %>%
  # Drop temporary columns
  select(grid_code, name, value) %>%
  ggplot() +
    geom_histogram(aes(value)) +
    theme_minimal() +
    labs(x = "N time grids are in the same group (after 1000 sbm)")

plotMyMatrix(coverage_subsample_cluster_mat, c(row = "grids", col = "grids"))

```

```{r}
estimateSimpleSBM(coverage_subsample_cluster_mat,model = "poisson",
                  estimOptions = list(plot = FALSE,
                                      verbosity = 0)) -> sbm_copartnership_coverage_subsample
plot(sbm_copartnership_coverage_subsample)
```

# OTU level simple subsampling

```{r}
subsample.equally.rowise <- function(x, n_subsample=5){
  S <- rownames(x)
  kept <- sample(S,size = n_subsample,replace = F)
  
  x[kept,]
}
```

```{r}
set.seed(1234)  
purrr::map(1:n_sim, \(x) subsample.equally.rowise(OTU_matrix_trucated, n_subsample = 500)) -> subsample_list
if(file.exists("outputs/heavy_computation_save/truncated_subsample_sbm_bact.Rdata")){

  load(file = "outputs/heavy_computation_save/truncated_subsample_sbm_bact.Rdata")

  }else{

  plan(multisession, workers = 10)
  subsample_list %>%
    furrr::future_map(\(x) x%>% 
                        as.matrix() %>%
                        log1p %>% 
                        estimateBipartiteSBM(netMat= ., model = "poisson", 
                                             dimLabels = c(row = "OTU", col = "grid_code"),
                                             estimOptions = list(plot = FALSE, verbosity = 0)),
                      .options = furrr_options(scheduling = 2),
                      .progress = T
  ) -> subsample_sbm_list

save(subsample_sbm_list, file = "outputs/heavy_computation_save/truncated_subsample_sbm_bact.Rdata")
  }
```

```{r}
map(1:n_sim, \(x) data.frame(subsample = subsample_sbm_list[[x]]$memberships$grid_code,
                             grid_code = colnames(subsample_list[[1]]),
                             sim = x)
    ) %>%
  bind_rows() %>%
  pivot_wider(names_from = sim, names_glue = "simulation{sim}", values_from = "subsample") %>%
  column_to_rownames("grid_code") -> cluster_subsample_df
```

```{r}
n_echantillons <- nrow(cluster_subsample_df)
subsample_cluster_mat <- matrix(0, nrow = n_echantillons, ncol = n_echantillons)

for (sim in 1:n_sim) {
  subsample_cluster_mat <- subsample_cluster_mat + outer(cluster_subsample_df[[sim]], cluster_subsample_df[[sim]], "==")
}

rownames(subsample_cluster_mat) <- rownames(cluster_subsample_df)
colnames(subsample_cluster_mat) <- rownames(cluster_subsample_df)
subsample_cluster_mat %>%
  as.data.frame() %>%
  rownames_to_column("grid_code") %>%
  pivot_longer(-grid_code, names_to = "name", values_to = "value") %>%
  # Create two temporary columns where the names are ALWAYS sorted
  # pmin/pmax ensures that (Sample_A, Sample_B) and (Sample_B, Sample_A) 
  # both become (Sample_A, Sample_B)
  mutate(
    p1 = pmin(grid_code, name),
    p2 = pmax(grid_code, name)
  ) %>%
  # Distinct now recognizes these rows as identical
  distinct(p1, p2, .keep_all = TRUE) %>%
  # Drop temporary columns
  select(grid_code, name, value) %>%
  ggplot() +
    geom_histogram(aes(value)) +
    theme_minimal() +
    labs(x = "N time grids are in the same group (after 1000 sbm)")

plotMyMatrix(subsample_cluster_mat, c(row = "grids", col = "grids"))

```

There are cleare group of sample that are appearing. 
It could be interesting to see what are teh clsuter found by an other sbm on this matrix

```{r}
estimateSimpleSBM(subsample_cluster_mat,model = "poisson",
                  estimOptions = list(plot = FALSE,
                                      verbosity = 0)) -> sbm_copartnership_subsample
plot(sbm_copartnership_subsample)
```

# OTU level subsampling with covariate

```{r}
library_size_mat = matrix(rep(colSums(OTU_matrix_trucated), 500), nrow = 500, ncol = 43, byrow = T)

library_size_mat[1,] == colSums(OTU_matrix_trucated)

if(file.exists("outputs/heavy_computation_save/subsample_covar_sbm_bact.Rdata")){

  load(file = "outputs/heavy_computation_save/subsample_covar_sbm_bact.Rdata")

  }else{

  plan(multisession, workers = 10)
  subsample_list %>%
    furrr::future_map(\(x) x%>% 
                        as.matrix() %>%
                        log1p %>% 
                        estimateBipartiteSBM(netMat= ., model = "poisson",
                                             dimLabels = c(row = "OTU", col = "grid_code"),
                                             estimOptions = list(plot = FALSE,
                                                                 nbCores =1,
                                                                 verbosity = 0),
                                             covariates   = list(library_size = library_size_mat)),
                      .options = furrr_options(scheduling = Inf),
                      .progress = T
  ) -> subsample_covar_sbm_list

save(subsample_covar_sbm_list, file = "outputs/heavy_computation_save/subsample_covar_sbm_bact.Rdata")
  }
```

```{r}
map(1:n_sim, \(x) data.frame(subsample = subsample_covar_sbm_list[[x]]$memberships$grid_code,
                             grid_code = colnames(subsample_list[[1]]),
                             sim = x)
    ) %>%
  bind_rows() %>%
  pivot_wider(names_from = sim, names_glue = "simulation{sim}", values_from = "subsample") %>%
  column_to_rownames("grid_code") -> cluster_subsample_covar_df
```

```{r}
n_echantillons <- nrow(cluster_subsample_covar_df)
subsample_covar_cluster_mat <- matrix(0, nrow = n_echantillons, ncol = n_echantillons)

for (sim in 1:n_sim) {
  subsample_covar_cluster_mat <- subsample_covar_cluster_mat + outer(cluster_subsample_covar_df[[sim]], cluster_subsample_covar_df[[sim]], "==")
}

rownames(subsample_covar_cluster_mat) <- rownames(cluster_subsample_covar_df)
colnames(subsample_covar_cluster_mat) <- rownames(cluster_subsample_covar_df)
subsample_covar_cluster_mat %>%
  as.data.frame() %>%
  rownames_to_column("grid_code") %>%
  pivot_longer(-grid_code, names_to = "name", values_to = "value") %>%
  # Create two temporary columns where the names are ALWAYS sorted
  # pmin/pmax ensures that (Sample_A, Sample_B) and (Sample_B, Sample_A) 
  # both become (Sample_A, Sample_B)
  mutate(
    p1 = pmin(grid_code, name),
    p2 = pmax(grid_code, name)
  ) %>%
  # Distinct now recognizes these rows as identical
  distinct(p1, p2, .keep_all = TRUE) %>%
  # Drop temporary columns
  select(grid_code, name, value) %>%
  ggplot() +
    geom_histogram(aes(value)) +
    theme_minimal() +
    labs(x = "N time grids are in the same group (after 1000 sbm)")

plotMyMatrix(subsample_covar_cluster_mat, c(row = "grids", col = "grids"))
```

There are clear group of sample that are appearing. 
It could be interesting to see what are the cluster found by an other sbm on this matrix

```{r}
estimateSimpleSBM(subsample_covar_cluster_mat,model = "poisson") -> sbm_copartnership_subsample_covar
plot(sbm_copartnership_subsample_covar)
```


```{r}

B <- as.data.frame(
    table(
      colnames(OTU_matrix_trucated),
      sbm_copartnership_subsample$memberships,
      sbm_copartnership_subsample_covar$memberships,
      sbm_copartnership_coverage_subsample$memberships

      )
  )

colnames(B) =  c( "grid_code", "subsample", "subsample_covar", "subsample_coverage", "Freq")

data.frame(library_size = sort(colSums(OTU_matrix_trucated), decreasing = TRUE),
           col_library_size = viridis(43)) %>%
  rownames_to_column("grid_code")-> col_library_size_df


B %>%
  left_join(col_library_size_df, by = "grid_code") -> B


w   <- which(B$Freq != 0)
B <- B[w, ]

B$library_size <- factor(B$library_size, 
                         levels = sort(unique(B$library_size), decreasing = TRUE))
alluvial(B[,c(1,2,3,4)],
         freq = B$Freq,
         col = B$col_library_size,
         alpha = 0.7,
         cex = 1.5,
         cex.axis = 1.5,
         xw = 0.2)
```


# At genus level

```{r}
ps_bacteria_genus_grid <- ps_bacteria %>%
  tax_glom(taxrank = "genus") %>%
  merge_samples("grid")

ps_bacteria_genus_grid %>%
  otu_table() %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("genus_code") %>%
  pivot_longer(-genus_code, names_to = "grid_code", values_to = "reads") %>%
  mutate(reads = floor(reads/10)) %>%
  filter(reads > 0) -> trucated_bacteria_genus_grid

```


# At genus level - Coverage rarefaction and SBM

```{r}
trucated_bacteria_genus_grid %>%
  pivot_wider(names_from = "grid_code", values_from = "reads", values_fill = 0) %>%
  column_to_rownames("genus_code") -> OTU_matrix_trucated_genus

str(as.matrix(OTU_matrix_trucated_genus))
```

## Find samples sizes (at reference coverage) - genus

```{r}
trucated_bacteria_genus_grid %>%
  group_by(grid_code) %>%
  mutate(sample_size = sum(reads)) %>%
  filter(reads <= 4, reads != 0) %>%
  group_by(grid_code, reads, sample_size) %>%
  summarize(count = n(), .groups = "drop") %>%
  pivot_wider(names_from = reads, values_from = count, values_fill = 0) %>%
  rename(f1 = "1", f2 = "2", f3 = "3", f4 = "4") %>%
  mutate(f1_hat = ifelse(f3 * f4 > 0, 4 * f2^2 / (3 * f3) - f2 * f3 / (2 * f4), 4 * f2^2 / (3 * (f3 + 1)) - f2 * f3 / (2 * (f4 + 1))),
         coverage = calculate.coverage.f1(f1, f2, sample_size),
         coverage_hat = calculate.coverage.f1(f1_hat, f2, sample_size)) -> grid_coverage_genus_df

grid_coverage_genus_df
```

```{r}
iNext_object_genus <- OTU_matrix_trucated_genus %>%
  iNEXT.3D::iNEXT3D(q = c(0), datatype = "abundance", nboot = 0)

iNext_object_genus$TDiNextEst$coverage_based %>%
  filter(Order.q == 0 & Method == "Observed") %>%
  filter(SC == min(SC, na.rm = TRUE)) %>%
  pull(SC) -> min_coverage_genus

OTU_matrix_trucated_genus %>%
  estimate3D(level = min_coverage_genus, q = 0, nboot = 0) %>%
  select(Assemblage, m) -> required_sizes_genus
```

## Rarefaction at reference coverage - genus

```{r}
set.seed(2365)
1:n_sim %>%
  map(\(x) OTU_matrix_trucated_genus %>%
        coverage_rarefy_matrix(required_sizes = required_sizes_genus)) -> coverage_rarefy_genus_list
```

```{r}
coverage_rarefy_genus_list[[1]] %>% str
coverage_rarefy_genus_list[[2]] %>% str
```

```{r}
if (file.exists("outputs/heavy_computation_save/sbm_coverage_bact_genus_list.Rdata")) {

  load(file = "outputs/heavy_computation_save/sbm_coverage_bact_genus_list.Rdata")

} else {

  plan(multisession, workers = 10)
  coverage_rarefy_genus_list %>%
    furrr::future_map(\(x) x %>%
                        as.matrix() %>%
                        log1p %>%
                        estimateBipartiteSBM(netMat = .,
                                             model = "poisson",
                                             dimLabels = c(row = "genus_code", col = "grid_code"),
                                             estimOptions = list(plot = FALSE,
                                                                 verbosity = 0,
                                                                  nbCores = 1)),
                      .options = furrr_options(scheduling = 2),
                      .progress = T
    ) -> coverage_subsample_genus_sbm_list

  save(coverage_subsample_genus_sbm_list, file = "outputs/heavy_computation_save/sbm_coverage_bact_genus_list.Rdata")
}
```

```{r}
map(1:n_sim, \(x) data.frame(subsample = coverage_subsample_genus_sbm_list[[x]]$memberships$grid_code,
                              grid_code = colnames(coverage_rarefy_genus_list[[1]]),
                              sim = x)
) %>%
  bind_rows() %>%
  pivot_wider(names_from = sim, names_glue = "simulation{sim}", values_from = "subsample") %>%
  column_to_rownames("grid_code") -> cluster_coverage_subsample_genus_df
```

```{r}
n_echantillons_genus <- nrow(cluster_coverage_subsample_genus_df)
coverage_subsample_cluster_genus_mat <- matrix(0, nrow = n_echantillons_genus, ncol = n_echantillons_genus)

for (sim in 1:n_sim) {
  coverage_subsample_cluster_genus_mat <- coverage_subsample_cluster_genus_mat + outer(cluster_coverage_subsample_genus_df[[sim]], cluster_coverage_subsample_genus_df[[sim]], "==")
}

rownames(coverage_subsample_cluster_genus_mat) <- rownames(cluster_coverage_subsample_genus_df)
colnames(coverage_subsample_cluster_genus_mat) <- rownames(cluster_coverage_subsample_genus_df)

coverage_subsample_cluster_genus_mat %>%
  as.data.frame() %>%
  rownames_to_column("grid_code") %>%
  pivot_longer(-grid_code, names_to = "name", values_to = "value") %>%
  mutate(
    p1 = pmin(grid_code, name),
    p2 = pmax(grid_code, name)
  ) %>%
  distinct(p1, p2, .keep_all = TRUE) %>%
  select(grid_code, name, value) %>%
  ggplot() +
    geom_histogram(aes(value)) +
    theme_minimal() +
    labs(x = "N time grids are in the same group (after 1000 sbm) - genus coverage")

plotMyMatrix(coverage_subsample_cluster_genus_mat, c(row = "grids", col = "grids"))
```

```{r}
estimateSimpleSBM(coverage_subsample_cluster_genus_mat, model = "poisson",
                  estimOptions = list(plot = FALSE,
                                      verbosity = 0)) -> sbm_copartnership_coverage_subsample_genus
plot(sbm_copartnership_coverage_subsample_genus)
```

# Genus level simple subsampling

```{r}
set.seed(1234)
purrr::map(1:n_sim, \(x) subsample.equally.rowise(OTU_matrix_trucated_genus, n_subsample = 500)) -> subsample_genus_list

if (file.exists("outputs/heavy_computation_save/truncated_subsample_sbm_bact_genus.Rdata")) {

  load(file = "outputs/heavy_computation_save/truncated_subsample_sbm_bact_genus.Rdata")

} else {

  plan(multisession, workers = 10)
  subsample_genus_list %>%
    furrr::future_map(\(x) x %>%
                        as.matrix() %>%
                        log1p %>%
                        estimateBipartiteSBM(netMat = ., model = "poisson",
                                             dimLabels = c(row = "genus_code", col = "grid_code"),
                                             estimOptions = list(plot = FALSE, 
                                                                  verbosity = 0,
                                                                  nbCores = 1)),
                      .options = furrr_options(scheduling = 2),
                      .progress = T
    ) -> subsample_genus_sbm_list

  save(subsample_genus_sbm_list, file = "outputs/heavy_computation_save/truncated_subsample_sbm_bact_genus.Rdata")
}
```

```{r}
map(1:n_sim, \(x) data.frame(subsample = subsample_genus_sbm_list[[x]]$memberships$grid_code,
                              grid_code = colnames(subsample_genus_list[[1]]),
                              sim = x)
) %>%
  bind_rows() %>%
  pivot_wider(names_from = sim, names_glue = "simulation{sim}", values_from = "subsample") %>%
  column_to_rownames("grid_code") -> cluster_subsample_genus_df
```

```{r}
n_echantillons_genus <- nrow(cluster_subsample_genus_df)
subsample_cluster_genus_mat <- matrix(0, nrow = n_echantillons_genus, ncol = n_echantillons_genus)

for (sim in 1:n_sim) {
  subsample_cluster_genus_mat <- subsample_cluster_genus_mat + outer(cluster_subsample_genus_df[[sim]], cluster_subsample_genus_df[[sim]], "==")
}

rownames(subsample_cluster_genus_mat) <- rownames(cluster_subsample_genus_df)
colnames(subsample_cluster_genus_mat) <- rownames(cluster_subsample_genus_df)

subsample_cluster_genus_mat %>%
  as.data.frame() %>%
  rownames_to_column("grid_code") %>%
  pivot_longer(-grid_code, names_to = "name", values_to = "value") %>%
  mutate(
    p1 = pmin(grid_code, name),
    p2 = pmax(grid_code, name)
  ) %>%
  distinct(p1, p2, .keep_all = TRUE) %>%
  select(grid_code, name, value) %>%
  ggplot() +
    geom_histogram(aes(value)) +
    theme_minimal() +
    labs(x = "N time grids are in the same group (after 1000 sbm) - genus subsample")

plotMyMatrix(subsample_cluster_genus_mat, c(row = "grids", col = "grids"))
```

There are clear groups of samples that are appearing.
It could be interesting to see what are the clusters found by another sbm on this matrix

```{r}
estimateSimpleSBM(subsample_cluster_genus_mat, model = "poisson",
                  estimOptions = list(plot = FALSE,
                                      verbosity = 0)) -> sbm_copartnership_subsample_genus
plot(sbm_copartnership_subsample_genus)
```

# Genus level subsampling with covariate

```{r}
library_size_genus_mat <- matrix(rep(colSums(OTU_matrix_trucated_genus), 500), nrow = 500, ncol = ncol(OTU_matrix_trucated_genus), byrow = T)

library_size_genus_mat[1, ] == colSums(OTU_matrix_trucated_genus)

if (file.exists("outputs/heavy_computation_save/subsample_covar_sbm_bact_genus.Rdata")) {

  load(file = "outputs/heavy_computation_save/subsample_covar_sbm_bact_genus.Rdata")

} else {

  plan(multisession, workers = 10)
  subsample_genus_list %>%
    furrr::future_map(\(x) x %>%
                        as.matrix() %>%
                        log1p %>%
                        estimateBipartiteSBM(netMat = ., model = "poisson",
                                             dimLabels = c(row = "genus_code", col = "grid_code"),
                                             estimOptions = list(plot = FALSE,
                                                                 nbCores = 1,
                                                                 verbosity = 0),
                                             covariates = list(library_size = library_size_genus_mat)),
                      .options = furrr_options(scheduling = Inf),
                      .progress = T
    ) -> subsample_covar_genus_sbm_list

  save(subsample_covar_genus_sbm_list, file = "outputs/heavy_computation_save/subsample_covar_sbm_bact_genus.Rdata")
}
```

```{r}
map(1:n_sim, \(x) data.frame(subsample = subsample_covar_genus_sbm_list[[x]]$memberships$grid_code,
                              grid_code = colnames(subsample_genus_list[[1]]),
                              sim = x)
) %>%
  bind_rows() %>%
  pivot_wider(names_from = sim, names_glue = "simulation{sim}", values_from = "subsample") %>%
  column_to_rownames("grid_code") -> cluster_subsample_covar_genus_df
```

```{r}
n_echantillons_genus <- nrow(cluster_subsample_covar_genus_df)
subsample_covar_cluster_genus_mat <- matrix(0, nrow = n_echantillons_genus, ncol = n_echantillons_genus)

for (sim in 1:n_sim) {
  subsample_covar_cluster_genus_mat <- subsample_covar_cluster_genus_mat + outer(cluster_subsample_covar_genus_df[[sim]], cluster_subsample_covar_genus_df[[sim]], "==")
}

rownames(subsample_covar_cluster_genus_mat) <- rownames(cluster_subsample_covar_genus_df)
colnames(subsample_covar_cluster_genus_mat) <- rownames(cluster_subsample_covar_genus_df)

subsample_covar_cluster_genus_mat %>%
  as.data.frame() %>%
  rownames_to_column("grid_code") %>%
  pivot_longer(-grid_code, names_to = "name", values_to = "value") %>%
  mutate(
    p1 = pmin(grid_code, name),
    p2 = pmax(grid_code, name)
  ) %>%
  distinct(p1, p2, .keep_all = TRUE) %>%
  select(grid_code, name, value) %>%
  ggplot() +
    geom_histogram(aes(value)) +
    theme_minimal() +
    labs(x = "N time grids are in the same group (after 1000 sbm) - genus subsample covar")

plotMyMatrix(subsample_covar_cluster_genus_mat, c(row = "grids", col = "grids"))
```

There are clear groups of samples that are appearing.
It could be interesting to see what are the clusters found by another sbm on this matrix

```{r}
estimateSimpleSBM(subsample_covar_cluster_genus_mat, model = "poisson") -> sbm_copartnership_subsample_covar_genus
plot(sbm_copartnership_subsample_covar_genus)
```

# Alluvial - genus level

```{r}
B_genus <- as.data.frame(
  table(
    colnames(OTU_matrix_trucated_genus),
    sbm_copartnership_subsample_genus$memberships,
    sbm_copartnership_subsample_covar_genus$memberships,
    sbm_copartnership_coverage_subsample_genus$memberships
  )
)

colnames(B_genus) <- c("grid_code", "subsample_genus", "subsample_covar_genus", "subsample_coverage_genus", "Freq")

data.frame(library_size = sort(colSums(OTU_matrix_trucated_genus), decreasing = TRUE),
           col_library_size = viridis(ncol(OTU_matrix_trucated_genus))) %>%
  rownames_to_column("grid_code") -> col_library_size_genus_df

B_genus %>%
  left_join(col_library_size_genus_df, by = "grid_code") -> B_genus

w_genus <- which(B_genus$Freq != 0)
B_genus <- B_genus[w_genus, ]

B_genus$library_size <- factor(B_genus$library_size,
                                levels = sort(unique(B_genus$library_size), decreasing = TRUE))

alluvial(B_genus[, c(6, 2, 3, 4)],
         freq = B_genus$Freq,
         col = B_genus$col_library_size,
         alpha = 0.7,
         cex = 1.5,
         cex.axis = 1.5,
         xw = 0.2)
```



```{r}
B_cross <- as.data.frame(
  table(
    colnames(OTU_matrix_trucated),
    sbm_copartnership_subsample$memberships,
    sbm_copartnership_subsample_genus$memberships,
    sbm_copartnership_subsample_covar$memberships,
    sbm_copartnership_subsample_covar_genus$memberships,
    sbm_copartnership_coverage_subsample$memberships,
    sbm_copartnership_coverage_subsample_genus$memberships
  )
)

colnames(B_cross) <- c("grid_code", "subsample_OTU", "subsample_genus",
                        "subsample_covar_OTU", "subsample_covar_genus",
                        "coverage_OTU", "coverage_genus", "Freq")

B_cross %>%
  left_join(col_library_size_df, by = "grid_code") -> B_cross

w_cross <- which(B_cross$Freq != 0)
B_cross <- B_cross[w_cross, ]

B_cross$library_size <- factor(B_cross$library_size,
                                levels = sort(unique(B_cross$library_size), decreasing = TRUE))

alluvial(B_cross[, c("library_size","subsample_OTU", "subsample_genus")],
         freq = B_cross$Freq,
         col = B_cross$col_library_size,
         alpha = 0.7,
         cex = 1.5,
         cex.axis = 1.5,
         xw = 0.2)

alluvial(B_cross[, c("library_size","subsample_covar_OTU", "subsample_covar_genus")],
         freq = B_cross$Freq,
         col = B_cross$col_library_size,
         alpha = 0.7,
         cex = 1.5,
         cex.axis = 1.5,
         xw = 0.2)
alluvial(B_cross[, c("library_size","coverage_OTU", "coverage_genus")],
         freq = B_cross$Freq,
         col = B_cross$col_library_size,
         alpha = 0.7,
         cex = 1.5,
         cex.axis = 1.5,
         xw = 0.2)
```







